---
title: "Notebook 2 — Pseudobulk DE & GO Analysis"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(edgeR)
library(ggplot2)
library(dplyr)
library(patchwork)
library(clusterProfiler)
library(org.Mm.eg.db)
library(sparseMatrixStats)
library(lme4)
library(lmerTest)

PROJ_DIR <- "/shared/projects/tp_2630_ubordeaux_neuromics_184418/projects/C14/"   # set this to your project path, e.g. "/shared/projects/.../projects/C10"
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "figures")
GIT_DIR  <- "~/cajal-project-alan"   # set this to your git checkout path, e.g. "~/cajal-project-alan"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

```{r}
#| label: load-data
sci_sub = readRDS(file.path(DATA_DIR, "sci-subset-atlas-annotated.rds"))
sci_sub$cell_type = sci_sub$layer2
```

---

## From composition change to cell-type-specific DE

In Notebook 1, you showed that spinal cord injury shifts the **composition** of cell types. But composition change is only one half of the story. The other half: **within each cell type, which genes change?**

A gene driving microglial activation is biologically distinct from one driving astrocyte reactivity — even if both appear as "upregulated in injured tissue" in a bulk experiment. Single-cell resolution lets us separate these signals.

The correct unit of replication is the biological replicate (animal), not the cell — testing gene by gene across individual cells inflates significance because cells from the same animal share correlated expression. `Libra::run_de` avoids this by aggregating counts per (cell type × replicate) into a **pseudobulk** profile, then running a bulk DE method (edgeR) on those profiles.

---

## 1. Pseudobulk DE on real SCI data

`Libra::run_de` runs pseudobulk edgeR per cell type. We run all three pairwise comparisons across all cell types and collect results into a single data frame with a `comparison` column.

```{r}
#| label: real-de-celltypes
#| eval: false
# Show cell type sizes across all conditions
# Hint: sort(table(sci_sub$cell_type), decreasing=TRUE)


```

```{r}
#| label: real-de
#| eval: false
# Define comparisons as a named list: young_vs_uninjured, young_vs_old, uninjured_vs_old
# For each comparison: subset sci_sub to those two conditions, run Libra::run_de,
#   add a $comparison column, rbind all results into de_res
# Add a $sig column (p_val_adj < 0.05 & |avg_logFC| > 1)


```

```{r}
#| label: real-de-volcanoes
#| eval: false
# 1. Bar chart: sig gene counts per cell type × comparison
#    aggregate(sig ~ cell_type + comparison, de_res, sum)
#    geom_col(position="dodge") + reorder cell_type by total sig genes
# 2. Identify top 6 cell types by sig count in young_vs_uninjured (store as top_cts)
# 3. Volcano plots for those 6 cell types (young_vs_uninjured only), wrap_plots(ncol=3)
# 4. Print top 10 hits for top_cts[1]


```

**Explore:** Which cell types respond most strongly to injury? Do the same cell types rank highly across all three comparisons, or does the pattern differ between young and old injury?

**Explore:** Which cell type shows the most DE genes across all three comparisons (young vs uninjured, young vs old, uninjured vs old)? Is the response cell-type-specific or broadly shared? What does this pattern tell you about the injury response relative to aging?

---

## 2. GO enrichment

Build the GO BP gene sets from `org.Mm.eg.db` once, then pass them directly to `enricher` and `GSEA` — this avoids re-querying the database on every call.

```{r}
#| label: go-build-sets
#| eval: false
# Build GO BP gene sets from org.Mm.eg.db once:
# go_ann = AnnotationDbi::select(org.Mm.eg.db, keys=..., columns=c("SYMBOL","GO","ONTOLOGY"), ...)
# Filter to ONTOLOGY == "BP", build go_t2g (term, gene) and go_t2n (term, name) tables


```

```{r}
#| label: go-analysis
#| eval: false
# Use top_cts[1] and comparison "young_vs_uninjured"
# sig_genes: upregulated genes (avg_logFC > 0 & p_val_adj < 0.05) for that ct × comparison
# enricher(gene=sig_genes, universe=all genes for that ct × comparison, TERM2GENE=go_t2g, TERM2NAME=go_t2n)
# dotplot(ego, showCategory=20)


```

**Explore:** What biological processes come up? Do they match your expectations for this cell type after injury?

---

## 3. GSEA

GO ORA only uses genes that pass a threshold. GSEA uses the **full ranked list** — every gene gets a fold change score, and the algorithm asks whether genes in a pathway cluster near the top or bottom. This is more sensitive and doesn't require an arbitrary cutoff.

```{r}
#| label: gsea
#| eval: false
# Filter de_res to top_cts[1] & "young_vs_uninjured"
# Build a named gene list: log2FC values sorted decreasing (all genes, no threshold)
# Hint: sort(setNames(de_ct$avg_logFC, de_ct$gene), decreasing=TRUE)
# GSEA(geneList=gene_list, TERM2GENE=go_t2g, TERM2NAME=go_t2n, minGSSize=15, maxGSSize=500)
# dotplot(gsea_res, showCategory=15, split=".sign") + facet_grid(. ~ .sign)


```

**Explore:** Do the GSEA results agree with GO? Are there pathways that GSEA picks up that ORA missed, or vice versa? What does the direction (activated vs suppressed) tell you about the biology?

**Explore:** Pick one enriched GO term from your results. Look up 2–3 of the genes driving it. Are any already known SCI or neuroinflammation markers? Does the unbiased GO analysis confirm or surprise you?

---

## 4. Custom gene sets — module scores

GO and GSEA test pre-defined databases of gene sets. Often you instead have a **custom gene panel** from a specific paper — a signature that isn't in GO/MSigDB at all. Seurat's `AddModuleScore` lets you score every cell for such a panel directly: for each cell, it computes the average expression of the panel genes minus the average expression of a control gene set (randomly sampled, matched for expression level), giving one number per cell instead of a per-gene test.

As an example, we use the **blood–brain barrier (BBB) dysfunction module** from Munji et al. (2019, *Nat Neurosci*) — a set of genes consistently induced in vascular endothelial cells across seven different CNS injury/disease models when the BBB breaks down.

```{r}
#| label: bbb-module-genes
#| eval: false
# Define the BBB dysfunction gene panel (Munji et al. 2019) as a character vector: bbb_genes
# Subset sci_sub to cell_type == "Vascular endothelial cells" -> vec
# NormalizeData(vec)
# Check how many bbb_genes are actually present in rownames(vec) (intersect)
# Run AddModuleScore(vec, features = list(<genes present>), name = "BBB_dysfunction")
# Pull vec@meta.data into vec_meta


```

```{r}
#| label: bbb-module-boxplot
#| eval: false
# Boxplot of BBB_dysfunction1 by label (condition), one box per condition, fill by label
# geom_boxplot(outlier.shape=NA) + geom_jitter(small size/alpha) for the individual cells


```

**Interpret:** Does the BBB dysfunction score look higher in injured (young/old) vs uninjured spinal cord? Is the spread within a condition mostly driven by a few animals, or evenly spread across cells?

Just like in section 1, cells from the same animal are not independent observations — a t-test or Wilcoxon test on cell-level module scores would treat every cell as a fresh data point and produce an overconfident p-value. The fix is the same: a **mixed model** with replicate as a random effect, so the test statistic reflects between-animal variation rather than between-cell variation.

```{r}
#| label: bbb-module-lmm
#| eval: false
# For each of the three comparisons (reuse the `comparisons` list from section 1):
#   subset vec_meta to those two conditions
#   fit lmerTest::lmer(BBB_dysfunction1 ~ label + (1 | replicate), data = <subset>)
#   pull the label coefficient's estimate and p-value from summary(fit)$coefficients
# Collect results (comparison, estimate, p_value) into one data frame and print it


```

**Explore:** Which comparison(s) show a significant difference in BBB dysfunction score? Does the direction (higher after injury?) match what you'd expect biologically? Would you have reached the same conclusion with a naive t-test on all cells pooled together?

---

::: {.callout-warning}
## CHECKPOINT
Which cell type has the most DE genes? Are the responses cell-type-specific or shared?

- What do GO and GSEA together suggest about what is happening in your chosen cell type after injury?
- Does the BBB dysfunction module score differ across the three comparisons, and does a mixed model change your conclusion relative to a naive cell-level test?
:::
