---
title: "Notebook 4 — Spatial Transcriptomics & Vespucci"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(Vespucci)
library(clusterProfiler)
library(org.Mm.eg.db)

PROJ_DIR <- "/shared/projects/tp_2630_ubordeaux_neuromics_184418/projects/C14/"   # set this to your project path, e.g. "/shared/projects/.../projects/C10"
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "figures")
GIT_DIR  <- "~/cajal-project-alan"   # set this to your git checkout path, e.g. "~/cajal-project-alan"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

---

Single-cell RNA-seq tells you which genes change in which cell types, but it discards the one thing a tissue has that a dissociated suspension does not: **location**. A neuron in the dorsal horn and one in the ventral horn can express identical genes yet have completely different function and connectivity. Spatial transcriptomics preserves the tissue coordinate system, letting you ask *where* in the spinal cord the injury response is concentrated.

This notebook works directly with our in-house SCI Visium data: first visualising raw spatial gene expression and RCTD cell type deconvolution, then applying Vespucci to map the spatial extent of the injury response and find spatially patterned DE genes.

---

## 1. In-house SCI Visium data

Our SCI Visium data is stored as a standard Seurat object with spatial coordinates in the metadata (`x`, `y`) rather than in the Seurat image slot. We therefore use `ggplot2` directly for all spatial plots.

```{r}
#| label: load-sci-visium
#| eval: false
#| fig-width: 14
#| fig-height: 5
sci_spatial = readRDS(file.path(DATA_DIR, "sci-spatial.rds"))
cat("Spots:", ncol(sci_spatial), "  Genes:", nrow(sci_spatial), "\n")
cat("Conditions:", paste(unique(sci_spatial$label), collapse = ", "), "\n")

ggplot(sci_spatial@meta.data, aes(x = x, y = y, colour = label)) +
    geom_point(size = 0.6) +
    scale_y_reverse() +
    coord_fixed() +
    facet_wrap(~ label) +
    theme_void() +
    labs(title = "SCI Visium — spot layout by condition") +
    theme(legend.position = "none")
```

```{r}
#| label: sci-normalise
#| eval: false
sci_spatial = NormalizeData(sci_spatial, assay = "RNA", verbose = FALSE)
```

```{r}
#| label: sci-genes-raw
#| eval: false
#| fig-width: 14
#| fig-height: 8
genes = c("Gfap", "Rbfox3")
expr_mat = as.matrix(GetAssayData(sci_spatial, assay = "RNA", layer = "data")[genes, ])
plot_df = do.call(rbind, lapply(genes, function(g) {
    cbind(sci_spatial@meta.data[, c("x", "y", "label")], expr = expr_mat[g, ], gene = g)
}))

ggplot(plot_df, aes(x = x, y = y, colour = expr)) +
    geom_point(size = 0.4) +
    scale_colour_gradientn(colours = c("grey90", "steelblue", "navy")) +
    scale_y_reverse() +
    coord_fixed() +
    facet_grid(gene ~ label) +
    theme_void() +
    labs(title = "Raw spot expression", colour = "log norm") +
    theme(strip.text = element_text(size = 8))
```

Raw spots are noisy — each spot captures 1–10 cells and the expression is sparse. We can reveal the underlying spatial pattern by fitting a local regression (LOESS) surface across the tissue and predicting onto a regular grid.

```{r}
#| label: sci-genes-loess
#| eval: false
#| fig-width: 14
#| fig-height: 8
meta = sci_spatial@meta.data
fixed_coords = expand.grid(x = seq(-400, 400, 5), y = seq(-150, 150, 5))

plot_list = list()
for (gene in genes) {
    meta$expr = as.numeric(GetAssayData(sci_spatial, assay = "RNA", layer = "data")[gene, ])
    conditions = unique(meta$label)

    coords = map_dfr(conditions, ~ {
        condition = .x
        fit = loess(expr ~ x * y, data = filter(meta, label == condition), span = 0.02, degree = 1)
        coords = tidyr::crossing(x = seq(-400, 400, 5), y = seq(-150, 150, 5)) %>% mutate(interp = predict(fit, newdata = .)) %>% drop_na(interp) %>%
        mutate(condition = condition)
    })

    range = quantile(coords$interp, c(0.01, 0.99))
    coords$interp = pmax(pmin(coords$interp, range[2]), range[1])
    brks = c(range[1] + 0.1 * diff(range), range[2] - 0.1 * diff(range))
    labels = c("min", "max")

    coords$gene = gene
    p = ggplot(coords, aes(x = x, y = y, fill = interp)) +
        facet_grid(gene ~ condition, switch = "y") +
        geom_raster() +
        scale_y_reverse(expand = c(0, 0), breaks = c(-150, 150), labels = c(expression("R" %->% ""), expression("" %<-% "L")),name = gene) +
        scale_x_continuous(expand = c(0, 0), breaks = c(-400, 400), labels = c(expression("" %<-% "Rostral"), expression("Caudal" %->% ""))) +
        scale_fill_distiller(name = "Expression", palette = "YlOrRd", direction = 1, limits = range, breaks = brks, labels = labels) +
        guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
        coord_fixed() +
        theme_bw() +
        theme(panel.grid = element_blank()) +
        theme(
            axis.title.x = element_blank(),
            axis.title.y = element_blank(),
            axis.text.y = element_text(angle = 90, hjust = c(1, 0)),
            axis.text.x = element_text(hjust = c(0, 1)),
            axis.ticks.x = element_blank(),
            axis.ticks.y = element_blank(),
            axis.ticks.length.x = unit(0, "lines"),
            axis.ticks.length.y = unit(0, "lines"),
            legend.position = "right",
            legend.justification = "bottom",
            legend.key.width = unit(0.18, "lines"),
            legend.key.height = unit(0.18, "lines")
        )
    if (gene != genes[1]) p = p + theme(strip.text.x = element_blank())
    if (gene != genes[length(genes)]) p = p + theme(legend.position = "none")
    plot_list[[gene]] = p
}
wrap_plots(plot_list, ncol = 1)
```

**Interpret:** Which genes show the sharpest spatial boundary in the interpolated surface? Does the pattern shift between conditions? What cell types do you think each gene marks, and what does this tell you about the anatomy of the spinal cord sections?

---

## 2. Cell type deconvolution — RCTD

Each Visium spot is ~55 µm in diameter and captures RNA from 1–10 cells. The measured expression is a mixture of whatever cell types happen to sit under that spot. Before interpreting spatial gene expression, we need to know *which cell types are where*.

**RCTD** (Robust Cell Type Decomposition, Cable et al. 2022) solves this by fitting a Poisson model to each spot's counts using our annotated scRNA-seq atlas as a reference. The `spacexr` package is not installed in the course environment so RCTD was run offline; we load the pre-computed proportions here.

```{r}
#| label: rctd-load
#| eval: false
rctd_res = readRDS(file.path(DATA_DIR, "sci-spatial-rctd.rds"))
# Long format: barcode × cell_type × weight × resolution
head(rctd_res)

# Join spatial coordinates
meta_coords = sci_spatial@meta.data[, c("barcode", "x", "y", "label")]
rctd_res = merge(rctd_res, meta_coords, by = "barcode")
```

```{r}
#| label: rctd-plot
#| eval: false
#| fig-width: 14
#| fig-height: 5
# Plot Immune cells (layer1) as a LOESS-smoothed spatial map
immune = rctd_res[rctd_res$cell_type == "Immune cells" & rctd_res$resolution == "layer1", ]

plot_df_rctd = map_dfr(unique(immune$label), ~ {
    condition = .x
    fit = loess(weight ~ x * y, data = filter(immune, label == condition), span = 0.02, degree = 1)
    tidyr::crossing(x = seq(-400, 400, 5), y = seq(-150, 150, 5)) %>%
        mutate(interp = predict(fit, newdata = .)) %>%
        drop_na(interp) %>%
        mutate(condition = condition)
})

range_rctd = quantile(plot_df_rctd$interp, c(0.01, 0.99))
plot_df_rctd$interp = pmax(pmin(plot_df_rctd$interp, range_rctd[2]), range_rctd[1])
brks_rctd = c(range_rctd[1] + 0.1 * diff(range_rctd), range_rctd[2] - 0.1 * diff(range_rctd))

ggplot(plot_df_rctd, aes(x = x, y = y, fill = interp)) +
    geom_raster() +
    scale_y_reverse(expand = c(0, 0), breaks = c(-150, 150), labels = c(expression("R" %->% ""), expression("" %<-% "L"))) +
    scale_x_continuous(expand = c(0, 0), breaks = c(-400, 400), labels = c(expression("" %<-% "Rostral"), expression("Caudal" %->% ""))) +
    scale_fill_distiller(name = "Weight", palette = "Greys", direction = 1, limits = range_rctd, breaks = brks_rctd, labels = c("min", "max")) +
    guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
    coord_fixed() +
    facet_wrap(~ condition) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text.y = element_text(angle = 90, hjust = c(1, 0)),
          axis.text.x = element_text(hjust = c(0, 1)),
          axis.ticks = element_blank()) +
    labs(title = "Immune cells (RCTD layer1)")
```

**Interpret:** Do the RCTD cell type proportions align with the tissue anatomy — neurons in grey matter, oligodendrocytes in white matter? Does the injured section show a shift in immune cell accumulation compared to uninjured?

**Explore:** Do the RCTD cell type proportions match the known anatomy of spinal cord sections — neurons concentrated in grey matter, oligodendrocytes in white matter, microglia more diffuse? Which spots show the highest immune cell fraction, and does this align with the injury site?

---

## 3. Vespucci — spatial perturbation mapping on the SCI atlas

Moran's I measures spatial autocorrelation **within a single tissue section** — it detects genes that cluster in space within one condition. But genes that are differentially expressed between conditions only in a specific spatial region have no spatial structure when you examine each condition in isolation, so Moran's I is blind to condition-specific spatial patterns.

**Vespucci** solves this directly: for each spot, it estimates a local AUC — how well a classifier can separate the two conditions using only that spot's spatial neighbourhood — producing a continuous spatial map of *where* the injury response is concentrated. It then uses the predicted AUC as a continuous covariate in an NBGMM DE test (`find_spatial_de`) to find genes that covary with local perturbation intensity, with replicate included as a random effect.

```{r}
#| label: vespucci-sci
#| eval: false
#| fig-width: 14
#| fig-height: 6
sci_spatial = readRDS(file.path(DATA_DIR, "sci-spatial-mapped.rds"))
counts_sc = GetAssayData(sci_spatial, layer = "counts")
meta_sc = sci_spatial@meta.data

vespucci_res = Vespucci::calculate_auc_fast(
    input = counts_sc,
    meta = meta_sc,
    label_col = "label",
    cell_type_col = "barcode"
)
auc_scores = setNames(vespucci_res$AUC$auc, vespucci_res$AUC$cell_type)
meta_sc$auc_pred = auc_scores[match(meta_sc$barcode, names(auc_scores))]

fit_sc = loess(auc_pred ~ x * y, data = meta_sc, span = 0.015)
meta_sc$auc_smooth = predict(fit_sc, meta_sc)
rng_sc = range(meta_sc$auc_smooth, na.rm = TRUE)
brks_sc = c(rng_sc[1] + 0.1 * diff(rng_sc), rng_sc[2] - 0.1 * diff(rng_sc))

ggplot(meta_sc[!is.na(meta_sc$auc_smooth), ],
       aes(x = x, y = y, fill = auc_smooth, color = auc_smooth)) +
    geom_point(size = 0.3, shape = 21, stroke = 0) +
    scale_fill_distiller(name = "AUC", palette = "RdYlBu", direction = -1,limits = rng_sc, breaks = brks_sc, labels = format(rng_sc, digits = 2)) +
    scale_color_distiller(name = "AUC", palette = "RdYlBu", direction = -1, limits = rng_sc, breaks = brks_sc, labels = format(rng_sc, digits = 2)) +
    guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE),
           color = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
    scale_y_reverse() + coord_fixed() + facet_wrap(~ label) + theme_bw() +
    theme(panel.grid = element_blank(), axis.title = element_blank(),
          axis.text = element_blank(), axis.ticks = element_blank(),
          legend.position = "right", legend.justification = "bottom",
          legend.key.width = unit(0.18, "lines"), legend.key.height = unit(0.18, "lines")) +
    labs(title = "Vespucci -- spatial perturbation map (SCI)")
```

```{r}
#| label: vespucci-sci-de
#| eval: false
library(nebula)

auc_df_sc = data.frame(barcode = meta_sc$barcode, auc = meta_sc$auc_pred)

de_sc = Vespucci::find_spatial_de(
    input = counts_sc,
    meta = meta_sc,
    auc_df = auc_df_sc,
    barcode_col = "barcode",
    replicate_col = "replicate"
)
head(de_sc, 10)
```

```{r}
#| label: vespucci-sci-de-plot
#| eval: false
#| fig-width: 12
#| fig-height: 14
sci_spatial = NormalizeData(sci_spatial, assay = "RNA", verbose = FALSE)
top_de_sc = head(de_sc$feature[de_sc$p_val_adj < 0.05], 6)
meta_sc_plot = meta_sc[, c("x", "y", "label")]

x_seq_sc = seq(floor(min(meta_sc_plot$x)), ceiling(max(meta_sc_plot$x)), length.out = 80)
y_seq_sc = seq(floor(min(meta_sc_plot$y)), ceiling(max(meta_sc_plot$y)), length.out = 80)

norm_sc = as.matrix(GetAssayData(sci_spatial, assay = "RNA", layer = "data"))

de_sc_plots = lapply(top_de_sc, function(gene) {
    meta_sc_plot$expr = norm_sc[gene, rownames(meta_sc_plot)]
    plot_df = map_dfr(unique(meta_sc_plot$label), ~ {
        fit = loess(expr ~ x * y, data = filter(meta_sc_plot, label == .x),
                    span = 0.02, degree = 1)
        tidyr::crossing(x = x_seq_sc, y = y_seq_sc) %>%
            mutate(interp = predict(fit, newdata = .), label = .x) %>%
            drop_na(interp)
    })
    rng_g = quantile(plot_df$interp, c(0.01, 0.99), na.rm = TRUE)
    plot_df$interp = pmax(pmin(plot_df$interp, rng_g[2]), rng_g[1])
    brks_g = c(rng_g[1] + 0.1 * diff(rng_g), rng_g[2] - 0.1 * diff(rng_g))
    ggplot(plot_df, aes(x = x, y = y, fill = interp)) +
        ggtitle(gene) +
        geom_raster() +
        scale_fill_distiller(name = "Expression", palette = "YlOrRd", direction = 1,limits = rng_g, breaks = brks_g, labels = c("min", "max")) +
        guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
        scale_x_continuous(expand = c(0, 0)) + scale_y_reverse(expand = c(0, 0)) +
        coord_fixed() + theme_bw() + theme(panel.grid = element_blank(),
              axis.title = element_blank(), axis.text = element_blank(),
              axis.ticks = element_blank()) +
        facet_wrap(~ label)
})
wrap_plots(de_sc_plots, ncol = 2)
```

```{r}
#| label: vespucci-sci-clusters
#| eval: false
# Cluster DE genes by their spatial expression pattern
sig_genes = de_sc$feature[de_sc$p_val_adj < 0.05]

genes_cluster_df = Vespucci::find_de_clusters(
    input = norm_sc,
    meta = meta_sc,
    de_genes = sig_genes,
    barcode_col = "barcode",
    norm = FALSE
)
table(genes_cluster_df$cluster)
```

```{r}
#| label: vespucci-sci-cluster-plot
#| eval: false
#| fig-width: 12
#| fig-height: 10
# Spatial map: mean expression of each gene cluster per spot
cluster_ids = sort(unique(genes_cluster_df$cluster))

cluster_spatial = map_dfr(cluster_ids, function(cl) {
    cl_genes = genes_cluster_df$gene[genes_cluster_df$cluster == cl]
    mean_expr = colMeans(norm_sc[cl_genes, meta_sc$barcode, drop = FALSE])
    data.frame(barcode = names(mean_expr), mean_expr = mean_expr, cluster = paste0("Cluster ", cl))
}) %>% left_join(meta_sc[, c("barcode", "x", "y")], by = "barcode")

cluster_plots = lapply(unique(cluster_spatial$cluster), function(cl) {
    df = filter(cluster_spatial, cluster == cl)
    fit = loess(mean_expr ~ x * y, data = df, span = 0.015)
    df$smooth = predict(fit, df)
    rng = range(df$smooth, na.rm = TRUE)
    brks = c(rng[1] + 0.1 * diff(rng), rng[2] - 0.1 * diff(rng))
    ggplot(df[!is.na(df$smooth), ], aes(x = x, y = y, fill = smooth, color = smooth)) +
        ggtitle(cl) +
        geom_point(size = 0.3, shape = 21, stroke = 0) +
        scale_fill_distiller(name = "Mean expr", palette = "YlOrRd", direction = 1,limits = rng, breaks = brks, labels = c("low", "high")) +
        scale_color_distiller(name = "Mean expr", palette = "YlOrRd", direction = 1,limits = rng, breaks = brks, labels = c("low", "high")) +
        guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE),
               color = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
        scale_y_reverse() + coord_fixed() + theme_bw() +
        theme(panel.grid = element_blank(), axis.title = element_blank(),
              axis.text = element_blank(), axis.ticks = element_blank())
})
wrap_plots(cluster_plots)
```

```{r}
#| label: vespucci-sci-gsea
#| eval: false
#| fig-width: 12
#| fig-height: 8
gsea_results = lapply(cluster_ids, function(cl) {
    cl_genes = genes_cluster_df$gene[genes_cluster_df$cluster == cl]
    tryCatch(
        enrichGO(gene = cl_genes, OrgDb = org.Mm.eg.db, keyType = "SYMBOL", ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.05, qvalueCutoff = 0.1),
        error = function(e) NULL
    )
})
names(gsea_results) = paste0("Cluster ", cluster_ids)

# Top 5 terms per cluster
gsea_df = map_dfr(names(gsea_results), function(cl_name) {
    res = gsea_results[[cl_name]]
    if (!is.null(res) && nrow(res) > 0)
        head(as.data.frame(res), 5) %>% mutate(cluster = cl_name)
})

ggplot(gsea_df, aes(x = reorder(Description, -log10(p.adjust)), y = -log10(p.adjust), fill = cluster)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    facet_wrap(~ cluster, scales = "free_y") +
    scale_x_discrete(labels = function(x) str_wrap(x, 30)) +
    theme_bw() + theme(panel.grid.minor = element_blank(), axis.title.y = element_blank()) +
    labs(y = "-log10 adjusted p-value", title = "GO:BP enrichment per DE gene cluster")
```

---

::: {.callout-warning}
## CHECKPOINT

What does spatial location add that single-cell alone could not tell you?

- Looking at the Vespucci AUC map for the SCI data: is the injury response uniform across the spinal cord cross-section, or concentrated in specific anatomical regions? How does this relate to the RCTD cell type distribution?
:::
