---
title: "Notebook 0 — R Warm-up & the Seurat Object"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

This notebook is a hands-on tour, not an exercise — every chunk is filled in and runnable. Work through it top to bottom before Notebook 1. By the end you should be comfortable with basic R syntax and with pulling anything you need — counts, metadata, individual genes — out of a Seurat object.

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(ggplot2)
library(dplyr)

PROJ_DIR <- "/shared/projects/tp_2630_ubordeaux_neuromics_184418/projects/C14/"   # set this to your project path, e.g. "/shared/projects/.../projects/C10"
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "figures")
GIT_DIR  <- "~/cajal-project-alan"   # set this to your git checkout path, e.g. "~/cajal-project-alan"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

---

## 1. R warm-up

If you already know R, skim this section and move on — it exists so nobody is blocked by syntax in Notebook 1.

### 1a. Vectors and indexing

R's basic unit is the vector — there is no such thing as a scalar, `5` is just a vector of length 1.

```{r}
#| label: vectors
gene_lengths = c(Gfap = 2996, Mbp = 1041, Rbfox3 = 1476, Cx3cr1 = 1653)
gene_lengths

gene_lengths["Mbp"]          # index by name
gene_lengths[c(1, 3)]        # index by position
gene_lengths[gene_lengths > 1500]   # index by logical condition
names(gene_lengths)
mean(gene_lengths)
sort(gene_lengths, decreasing = TRUE)
```

### 1b. Data frames

A data frame is a table: columns can be different types, every column has the same length. This is exactly what Seurat uses for per-cell metadata.

```{r}
#| label: dataframes
toy = data.frame(
    gene = c("Gfap", "Mbp", "Rbfox3", "Cx3cr1"),
    cell_type = c("Astrocyte", "Oligodendrocyte", "Neuron", "Microglia"),
    mean_expr = c(3.2, 5.1, 1.8, 2.4)
)
toy
str(toy)          # structure: types + first values of each column
toy$mean_expr     # $ pulls out a single column as a vector
toy[toy$mean_expr > 2, ]          # row filter, base R
toy[, c("gene", "cell_type")]     # column selection, base R
```

The same two operations with `dplyr`, which you'll see throughout the course notebooks:

```{r}
#| label: dplyr-basics
toy %>% filter(mean_expr > 2)
toy %>% select(gene, cell_type)
toy %>% arrange(desc(mean_expr))
toy %>% group_by(cell_type) %>% summarise(mean_expr = mean(mean_expr), .groups = "drop")
```

`%>%` (or the base-R `|>`) is the pipe: `x %>% f()` is just `f(x)`. Read it as "take `x`, then do `f`".

### 1c. Functions

```{r}
#| label: functions
log2fc = function(mean_a, mean_b) {
    log2(mean_a / mean_b)
}
log2fc(8, 2)

# apply a function across a vector without a loop
sapply(gene_lengths, function(x) x / 1000)
```

### 1d. Common gotchas

```{r}
#| label: gotchas
# `=` and `<-` are (almost always) interchangeable for assignment
x = 1
y <- 2

# indexing starts at 1, not 0
letters[1]

# NA propagates — check for it explicitly
v = c(1, 2, NA, 4)
mean(v)                  # NA
mean(v, na.rm = TRUE)    # 2.333...
is.na(v)

# `==` compares, `<-`/`=` assigns — a classic typo source
1 == 1
```

> **Try it yourself:** Using `toy` above, write one pipe that keeps only genes with `mean_expr > 2`, sorts them descending by `mean_expr`, and selects just `gene` and `mean_expr`.

---

## 2. What is a Seurat object?

A `Seurat` object bundles everything about a single-cell experiment into one structure:

- **assays** — one or more expression matrices (e.g. `RNA`); each assay can hold multiple **layers** (`counts`, `data` = normalised, `scale.data` = scaled)
- **meta.data** — a data frame with one row per cell (QC metrics, condition, cluster labels, ...)
- **reductions** — dimensionality reductions (`pca`, `harmony`, `umap`), once computed
- **active.ident** — the "current" grouping used by default in many plotting functions

Genes are **rows** (features), cells are **columns** (samples) — the opposite convention from a typical data frame.

```{r}
#| label: load-object
sci = readRDS(file.path(DATA_DIR, "sci-subset-atlas-no-annotation.rds"))
class(sci)
sci
```

The print method above is the fastest way to sanity-check an object: assay name, number of features/samples, which layers exist. `slotNames()` shows every slot in the S4 structure if you want to go deeper:

```{r}
#| label: slots
slotNames(sci)
```

---

## 3. Object dimensions and identity

```{r}
#| label: dims
dim(sci)                # features x cells
nrow(sci)                # number of genes
ncol(sci)                # number of cells

# genes and cell barcodes
head(rownames(sci))      # == Features(sci)
head(colnames(sci))      # == Cells(sci)
```

`Idents(sci)` is the object's default grouping variable — often the cluster or condition used implicitly by plotting functions:

```{r}
#| label: idents
levels(Idents(sci))
table(Idents(sci))
```

---

## 4. The count matrix

`GetAssayData()` pulls out a layer as a matrix. Raw UMI counts live in the `"counts"` layer.

```{r}
#| label: get-counts
counts = GetAssayData(sci, layer = "counts")
class(counts)     # a sparse matrix (dgCMatrix), not a base R matrix
dim(counts)
counts[1:5, 1:4]  # rows = genes, columns = cells
```

Single-cell count matrices are >90% zeros, so Seurat stores them sparsely — only non-zero values are kept in memory. Converting to a dense `matrix` for the full object would blow up your RAM; only densify small subsets.

```{r}
#| label: sparsity
n_zero = sum(counts == 0)
1 - n_zero / length(counts)   # fraction of entries that are zero
```

Row/column sums give you the two most basic per-gene and per-cell summaries:

```{r}
#| label: sums
lib_size = colSums(counts)      # total UMI per cell ("library size")
gene_total = rowSums(counts)    # total UMI per gene, across all cells

head(sort(lib_size, decreasing = TRUE))
head(sort(gene_total, decreasing = TRUE))
```

---

## 5. Metadata

Per-cell metadata lives in `sci@meta.data` (equivalently `sci[[]]`), one row per cell, aligned to `colnames(sci)`.

```{r}
#| label: metadata-basics
head(sci@meta.data)
colnames(sci@meta.data)
str(sci@meta.data)
```

Individual columns can be pulled with `$` or `[[ ]]`, same as any data frame:

```{r}
#| label: metadata-columns
head(sci$label)              # condition: uninjured / young / old
head(sci$replicate)          # biological replicate (animal)
head(sci$nCount_RNA)         # total UMI per cell (Seurat computes this automatically)
head(sci$nFeature_RNA)       # number of genes detected per cell

table(sci$label)
table(sci$replicate)
```

You can summarise metadata with the same `dplyr` verbs as any table:

```{r}
#| label: metadata-summarise
sci@meta.data %>%
    group_by(label) %>%
    summarise(
        n_cells = n(),
        mean_umi = mean(nCount_RNA),
        mean_genes = mean(nFeature_RNA),
        .groups = "drop"
    )
```

New metadata columns are added the same way — `PercentageFeatureSet()` is a built-in helper for one common one, percentage of mitochondrial reads (mouse mito genes are prefixed `mt-`, e.g. `mt-Nd1`, `mt-Co1`):

```{r}
#| label: add-metadata
sci[["pct_mt"]] = PercentageFeatureSet(sci, pattern = "^mt-")
head(sci$pct_mt)

# equivalently, from the raw counts directly:
mito_genes = grep("^mt-", rownames(sci), value = TRUE)
head(mito_genes)
```

---

## 6. Looking at specific genes

First, check a gene is actually in the object — a typo or wrong species case (`GFAP` vs `Gfap`) is the most common silent failure here:

```{r}
#| label: gene-exists
"Gfap" %in% rownames(sci)
"GFAP" %in% rownames(sci)   # wrong case — this data is mouse, so gene symbols are Title Case
```

Pull one gene's raw counts by row-indexing the matrix — this returns a plain named vector, one value per cell:

```{r}
#| label: one-gene-counts
gfap_counts = counts["Gfap", ]
head(gfap_counts)
summary(gfap_counts)
mean(gfap_counts > 0)   # fraction of cells with any detected expression
```

`FetchData()` is usually more convenient: it returns a tidy data frame combining any mix of genes, metadata columns, and reduction coordinates by name.

```{r}
#| label: fetchdata
gene_df = FetchData(sci, vars = c("Gfap", "Mbp", "Rbfox3", "label"))
head(gene_df)

gene_df %>%
    group_by(label) %>%
    summarise(mean_Gfap = mean(Gfap), .groups = "drop")
```

`VlnPlot()` is the quickest way to visualise one or more genes across a grouping variable:

```{r}
#| label: vlnplot
VlnPlot(sci, features = c("Gfap", "Mbp", "Rbfox3"), group.by = "label", pt.size = 0)
```

---

## 7. Plotting with ggplot2

`VlnPlot()`, `DimPlot()`, etc. are convenience wrappers that build a `ggplot` object for you. For anything custom — a specific comparison, a non-standard layout — pull the numbers out into a plain data frame (with `FetchData()`, `colSums()`, or `@meta.data`) and plot it yourself. This is the pattern used throughout the course notebooks.

### 7a. Histogram from a vector

```{r}
#| label: ggplot-histogram
lib_df = data.frame(lib_size = lib_size)

ggplot(lib_df, aes(x = lib_size)) +
    geom_histogram(bins = 50, fill = "steelblue", colour = "white", alpha = 0.7) +
    scale_x_log10() +
    labs(title = "Library size per cell", x = "Total UMI (log scale)", y = "Number of cells") +
    theme_minimal()
```

### 7b. Boxplot / violin grouped by metadata

`sci` hasn't been normalised yet at this point (only the `counts` layer exists — see Section 2), so `gene_df$Gfap` is a raw UMI count; that's why we plot `Gfap + 1` on a log scale rather than a linear axis.

```{r}
#| label: ggplot-boxplot
gene_df %>%
    ggplot(aes(x = label, y = Gfap + 1, fill = label)) +
    geom_violin(alpha = 0.5) +
    geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
    scale_y_log10() +
    scale_fill_manual(values = c("steelblue", "tomato", "seagreen")) +
    labs(title = "Gfap raw counts by condition", x = "Condition", y = "Count + 1 (log scale)") +
    theme_minimal() +
    theme(legend.position = "none")
```

This is (up to styling) exactly what `VlnPlot(sci, features = "Gfap", group.by = "label")` draws for you — useful to know once you want a layout `VlnPlot()` doesn't support directly.

### 7c. Scatter plot of two genes

Most cells have zero counts for both genes, so a small amount of jitter keeps the origin from hiding how many points are actually there.

```{r}
#| label: ggplot-scatter
ggplot(gene_df, aes(x = Gfap, y = Mbp, colour = label)) +
    geom_jitter(width = 0.3, height = 0.3, alpha = 0.4, size = 0.8) +
    scale_colour_manual(values = c("steelblue", "tomato", "seagreen")) +
    labs(title = "Gfap vs Mbp raw counts", x = "Gfap (count)", y = "Mbp (count)", colour = "Condition") +
    theme_minimal()
```

### 7d. Bar plot from a summary table

```{r}
#| label: ggplot-barplot
sci@meta.data %>%
    group_by(label) %>%
    summarise(n_cells = n(), .groups = "drop") %>%
    ggplot(aes(x = label, y = n_cells, fill = label)) +
    geom_col() +
    scale_fill_manual(values = c("steelblue", "tomato", "seagreen")) +
    labs(title = "Cells per condition", x = "Condition", y = "Number of cells") +
    theme_minimal() +
    theme(legend.position = "none")
```

> **Try it yourself:** Adapt 7a to facet the library-size histogram by `replicate` (`facet_wrap(~ replicate)`) instead of pooling all cells together.

---

## 8. Subsetting the object

`subset()` filters cells (and optionally genes) while keeping the object's structure intact — metadata, counts, and (once computed) reductions all stay aligned.

```{r}
#| label: subset-cells
sci_uninj = subset(sci, subset = label == "uninjured")
dim(sci_uninj)
table(sci_uninj$label)
```

Square-bracket indexing works the same way as a matrix: `[genes, cells]`.

```{r}
#| label: subset-bracket
sci_two_genes = sci[c("Gfap", "Mbp"), ]
dim(sci_two_genes)

sci_first10 = sci[, 1:10]
dim(sci_first10)
```

---

## 9. Cheat sheet

| Want | Code |
|---|---|
| Dimensions (genes x cells) | `dim(sci)` |
| Gene names / cell barcodes | `rownames(sci)` / `colnames(sci)` |
| Raw counts matrix | `GetAssayData(sci, layer = "counts")` |
| Normalised data (after `NormalizeData()`) | `GetAssayData(sci, layer = "data")` |
| All metadata | `sci@meta.data` or `sci[[]]` |
| One metadata column | `sci$label` |
| One gene, all cells (vector) | `GetAssayData(sci, layer = "counts")["Gfap", ]` |
| Genes + metadata together (data frame) | `FetchData(sci, vars = c("Gfap", "label"))` |
| Filter cells by metadata | `subset(sci, subset = label == "uninjured")` |
| Filter genes/cells by position | `sci[genes, cells]` |

You now have the tools to explore the object however you like. Notebook 1 picks up from here: QC filtering, normalisation, integration, and cell type annotation on this same spinal cord injury dataset.
