---
title: "Notebook 1 — Statistical Inference & Differential Expression (Solutions)"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(harmony)
library(edgeR)
library(ggplot2)
library(ggalluvial)
library(dplyr)
library(patchwork)
library(clusterProfiler)
library(org.Mm.eg.db)
library(splatter)
library(sparseMatrixStats)

PROJ_DIR <- "~/cajal-project-alan/"
DATA_DIR <- "~/cajal-project-alan/data/"
FIG_DIR  <- "~/cajal-project-alan/figures/"
GIT_DIR  <- "~/cajal-project-alan/"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

---

## 1. Raw count distributions 

```{r}
#| label: load-raw
sci = readRDS(file.path(DATA_DIR, "sci-subset-atlas-no-annotation.rds"))
counts = GetAssayData(sci, layer = "counts")
```

```{r}
#| label: raw-distributions
p1 = ggplot(data.frame(x = colSums(counts)), aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "steelblue", colour = "white", alpha = 0.4) +
    scale_x_log10() +
    labs(title = "Library size per cell", x = "Total UMI count (log scale)", y = "Density") +
    theme_minimal()

set.seed(42)
sample_cells = sample(ncol(counts), 1000)
p2 = ggplot(data.frame(x = counts[, sample_cells]@x), aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "tomato", colour = "white", alpha = 0.4) +
    scale_x_log10() +
    labs(title = "All non-zero counts (100 random cells)", x = "UMI count (log scale)", y = "Density") +
    theme_minimal()

p1 + p2
```

---

## 2. Simulating count data

```{r}
#| label: simulate-counts
sim_poisson = simulate_counts_naive("poisson", n_genes = 500, n_cells = 100)
sim_gaussian = simulate_counts_naive("gaussian", n_genes = 500, n_cells = 100)
sim_nb = simulate_counts_naive("nb", n_genes = 500, n_cells = 100)

counts_df = rbind(
    data.frame(x = sim_gaussian[sim_gaussian > 0], model = "Gaussian"),
    data.frame(x = sim_poisson[sim_poisson > 0], model = "Poisson"),
    data.frame(x = sim_nb[sim_nb > 0], model = "NB")
)
counts_df$model = factor(counts_df$model, levels = c("Gaussian", "Poisson", "NB"))

ggplot(counts_df, aes(x = x)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", colour = "white", alpha = 0.4) +
    scale_x_log10() +
    facet_wrap(~ model, nrow = 1) +
    labs(x = "Count (log scale)", y = "Density") +
    theme_minimal()
```

---

## 3. Why NBGMM?

To model differential expression in scRNA-seq properly, we need a distribution that matches how count data actually behaves. This section builds intuition for why the **Negative Binomial Generalised Mixed Model (NBGMM)** is the right choice.

### 3a. Overdispersion — why NB over Poisson

**The Poisson assumption.** The simplest count model is Poisson, where the variance equals the mean:

$$\text{Var}[X] = \mathbb{E}[X] = \mu$$

In scRNA-seq, this would mean a gene with mean expression 10 should have variance 10 across cells. In practice, the variance is much higher — this is **overdispersion**.

**Why scRNA-seq is overdispersed.** Gene expression is inherently stochastic. Transcription happens in bursts; cells in the same population differ in their regulatory state, cell cycle phase, and local microenvironment. This biological cell-to-cell variation adds variance on top of the Poisson sampling noise.

**The Negative Binomial distribution** adds a dispersion parameter `size` (also written φ or 1/α):

$$\text{Var}[X] = \mu + \frac{\mu^2}{\text{size}}$$

As `size` → ∞, the NB approaches Poisson. Small `size` (e.g. 0.1–1) means high overdispersion — typical for scRNA-seq.

**What about zero-inflated models?** Single-cell data contains many zeros — often >90% of entries are zero. These zeros come from two sources:

- **Structural zeros**: the gene is genuinely not expressed in that cell
- **Dropout zeros**: the gene was expressed but not captured due to low sequencing depth or PCR failure

A **Zero-Inflated Negative Binomial (ZINB)** model explicitly models this:

$$P(X = 0) = \pi + (1 - \pi) \cdot P_\text{NB}(X = 0)$$

where π is the probability of a structural (dropout) zero. However, in practice NB is often preferred: modern 10x Chromium protocols have lower dropout rates than older protocols, and the two types of zeros are statistically hard to separate. NB already predicts many zeros naturally for low-mean genes, and adding the zero-inflation parameter often increases model complexity without improving DE results.

> **Predict before you run:** On a log-log plot of variance vs. mean, where would you expect Poisson data to fall relative to NB?

```{r}
#| label: demo-overdispersion
set.seed(42)
n_genes_mv = 500
n_cells_mv = 100
gene_means = exp(rnorm(n_genes_mv, mean = log(5), sd = 1.5))

sim_pois_mv = matrix(rpois(n_genes_mv * n_cells_mv, lambda = gene_means),
    nrow = n_genes_mv, ncol = n_cells_mv)
sim_nb_mv = matrix(rnbinom(n_genes_mv * n_cells_mv, size = 0.5, mu = gene_means),
    nrow = n_genes_mv, ncol = n_cells_mv)

real_means = sparseMatrixStats::rowMeans2(counts)
real_vars = sparseMatrixStats::rowSds(counts)^2

mv_df = rbind(
    data.frame(mean = rowMeans(sim_pois_mv), var = apply(sim_pois_mv, 1, var), model = "Poisson (simulated)"),
    data.frame(mean = rowMeans(sim_nb_mv), var = apply(sim_nb_mv, 1, var), model = "NB (simulated)"),
    data.frame(mean = real_means, var = real_vars, model = "Real data")
)
mv_df = mv_df[mv_df$mean > 0 & mv_df$var > 0, ]
mv_df$model = factor(mv_df$model, levels = c("Poisson (simulated)", "NB (simulated)", "Real data"))

ggplot(mv_df, aes(x = mean, y = var)) +
    geom_point(alpha = 0.4, size = 0.6, colour = "steelblue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "tomato") +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ model, nrow = 1, scales = "free") +
    labs(
        title = "Mean–variance relationship",
        subtitle = "Real scRNA-seq data is overdispersed like NB — Poisson is constrained to the diagonal",
        x = "Gene mean", y = "Gene variance"
    ) +
    theme_minimal()
```

### 3b. Replicate structure — why a mixed model

Each biological replicate (animal/sample) has its own baseline expression level — a **random effect** shared by all cells from that sample. When you pool cells across replicates and ignore this structure, the extra between-replicate variance inflates the apparent spread of each group, making it easy to call false positives.

The **mixed model** in NBGMM accounts for this: it explicitly models the within-replicate correlation, so that the test statistic reflects between-replicate variation (the true unit of replication) rather than between-cell variation.

> **Predict before you run:** If each replicate has a random shift in mean expression, what will the pooled distribution look like compared to each individual replicate?

```{r}
#| label: demo-replicates
#| fig-width: 14
#| fig-height: 6
set.seed(42)
n_samples_demo = 6
n_cells_demo = 80
base_mean = 5
de_fc_demo = 2

group_demo = c(rep("Control", 3), rep("Treated", 3))
group_fc = c(rep(1, 3), rep(de_fc_demo, 3))
sample_effects = exp(rnorm(n_samples_demo, mean = 0, sd = 0.5))

cells_list = lapply(seq_len(n_samples_demo), function(i) {
    mu_i = base_mean * sample_effects[i] * group_fc[i]
    data.frame(
        count = rnbinom(n_cells_demo, size = 1, mu = mu_i),
        replicate = paste0("S", i),
        group = group_demo[i]
    )
})
cells_df = do.call(rbind, cells_list)
cells_df$replicate = factor(cells_df$replicate)

p_pooled = ggplot(cells_df, aes(x = count + 1, fill = group, colour = group)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, position = "identity", alpha = 0.3, colour = "white") +
    geom_density(alpha = 0) +
    scale_x_log10() +
    scale_fill_manual(values = c("steelblue", "tomato")) +
    scale_colour_manual(values = c("steelblue", "tomato")) +
    labs(
        title = "Pooled (ignoring replicates)",
        subtitle = "Broad mixture — extra width is replicate noise,\nnot biology",
        x = "Count + 1 (log scale)", y = "Density", fill = "Group"
    ) +
    guides(colour = "none") +
    theme_minimal()

p_per_sample = ggplot(cells_df, aes(x = count + 1, fill = group, colour = group)) +
    geom_histogram(aes(y = after_stat(density)), bins = 25, colour = "white", alpha = 0.3) +
    geom_density(alpha = 0) +
    scale_x_log10() +
    scale_fill_manual(values = c("steelblue", "tomato")) +
    scale_colour_manual(values = c("steelblue", "tomato")) +
    facet_wrap(~ replicate, nrow = 2) +
    labs(
        title = "Per replicate",
        subtitle = "Each replicate follows a tight NB —\nbut means shift between replicates",
        x = "Count + 1 (log scale)", y = "Density", fill = "Group"
    ) +
    guides(colour = "none") +
    theme_minimal()

p_pooled + p_per_sample + plot_layout(widths = c(1, 2))
```

### 3c. DE effect size (`de_fc`)

```{r}
#| label: demo-de-fc
sim_low_fc = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 1.5, n_samples = 4)
sim_high_fc = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 4, n_samples = 4)

lfc_df = rbind(
    transform(compute_logfc(sim_low_fc), scenario = "de_fc = 1.5"),
    transform(compute_logfc(sim_high_fc), scenario = "de_fc = 4")
)

ggplot(lfc_df, aes(x = logFC, fill = is_de, colour = is_de)) +
    geom_histogram(aes(y = after_stat(density)), bins = 60, position = "identity", alpha = 0.3, colour = "white") +
    geom_density(alpha = 0) +
    facet_wrap(~ scenario, nrow = 1) +
    scale_fill_manual(values = c("grey70", "tomato"), labels = c("not DE", "true DE")) +
    scale_colour_manual(values = c("grey50", "tomato"), labels = c("not DE", "true DE")) +
    labs(x = "log2 fold change (group means)", y = "Density", fill = NULL, colour = NULL) +
    theme_minimal()
```

### 3d. DE probability (`de_prob`)

```{r}
#| label: demo-de-prob
sim_low_prob = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.05, de_fc = 2, n_samples = 4)
sim_high_prob = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.4, de_fc = 2, n_samples = 4)

de_df = rbind(
    data.frame(scenario = "de_prob = 0.05",
        type = c("DE", "not DE"),
        n = c(length(sim_low_prob$de_genes), 1000 - length(sim_low_prob$de_genes))),
    data.frame(scenario = "de_prob = 0.40",
        type = c("DE", "not DE"),
        n = c(length(sim_high_prob$de_genes), 1000 - length(sim_high_prob$de_genes)))
)

ggplot(de_df, aes(x = scenario, y = n, fill = type)) +
    geom_col() +
    scale_fill_manual(values = c("tomato", "grey70")) +
    labs(y = "Number of genes", fill = NULL) +
    theme_minimal()
```

### 3e. Sequencing depth between replicates (`replicate_effect`)

Real experiments often have systematic differences in sequencing depth across replicates — one sample may have been sequenced more deeply than another, or RNA quality varied. The `replicate_effect` parameter in `simulate_counts()` introduces this by downsampling each replicate's library to a different depth. Setting it to 0 gives perfectly equal depth; larger values exaggerate the variation.

> **Predict before you run:** How would systematic depth differences across replicates affect a DE test that does not account for library size?

```{r}
#| label: replicate-effect
sim_no_effect = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 2, n_samples = 4, replicate_effect = 0)
sim_strong_effect = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 2, n_samples = 4, replicate_effect = 0.2)

real_lib_df = data.frame(
    lib_size = colSums(counts),
    batch = sci@meta.data$replicate,
    scenario = "Real data"
)
median_order = real_lib_df %>%
    group_by(batch) %>%
    summarise(med = median(lib_size), .groups = "drop") %>%
    arrange(desc(med)) %>%
    pull(batch)

lib_df = rbind(
    data.frame(lib_size = colSums(sim_no_effect$counts), batch = sim_no_effect$metadata$replicate, scenario = "replicate_effect = 0"),
    data.frame(lib_size = colSums(sim_strong_effect$counts), batch = sim_strong_effect$metadata$replicate, scenario = "replicate_effect = 0.2"),
    real_lib_df
)
lib_df$scenario = factor(lib_df$scenario, levels = c("replicate_effect = 0", "replicate_effect = 0.2", "Real data"))
lib_df$batch = factor(lib_df$batch, levels = c(paste0("Batch", 1:4), as.character(median_order)))

ggplot(lib_df, aes(x = batch, y = lib_size)) +
    geom_boxplot(fill = "steelblue", alpha = 0.5, outlier.shape = NA) +
    scale_y_log10() +
    facet_wrap(~ scenario, nrow = 1, scales = "free") +
    labs(x = "Sample", y = "Library size (total UMI, log scale)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Interpret:** Does the real data show more variability in library size across replicates than the `replicate_effect = 0` simulation? What might cause this in practice?

---

## 4. QC & normalisation

```{r}
#| label: qc-metrics
#| message: false
sci[["pct_mt"]] = PercentageFeatureSet(sci, pattern = "^mt-")

VlnPlot(sci,
        features = c("nFeature_RNA", "nCount_RNA", "pct_mt"),
        group.by = "label",
        pt.size  = 0)
```

```{r}
#| label: qc-filter
sci = subset(sci, subset = nCount_RNA > 200 & nCount_RNA < 25000 & pct_mt < 20)
```

```{r}
#| label: normalise
#| message: false
sci = NormalizeData(sci)
sci = FindVariableFeatures(sci, nfeatures = 3000)
sci = ScaleData(sci)
sci = RunPCA(sci, npcs = 30)
```

---

## 5. Harmony integration & UMAP

```{r}
#| label: harmony-umap
#| fig-width: 14
#| fig-height: 7
sci = RunHarmony(sci, group.by.vars = "replicate")
sci = RunUMAP(sci, reduction = "harmony", dims = 1:20)
DimPlot(sci, group.by = c("label", "replicate"), ncol = 2) & coord_fixed()
```

---

## 6. Cell type annotation

```{r}
#| label: clustering
#| message: false
#| fig-width: 18
#| fig-height: 7
sci = FindNeighbors(sci, reduction = "harmony", dims = 1:20)

resolutions = c(0.2, 0.5, 1.0)
for (res in resolutions) {
    sci = FindClusters(sci, resolution = res, verbose = FALSE)
}

p_res_list = lapply(resolutions, function(res) {
    DimPlot(sci, group.by = paste0("RNA_snn_res.", res), label = TRUE, label.size = 3, repel = TRUE) +
        ggtitle(paste0("resolution = ", res)) +
        coord_fixed() +
        theme(legend.position = "none")
})
wrap_plots(p_res_list, nrow = 1)
```

```{r}
#| label: load-annotation-clusters
#| fig-width: 14
#| fig-height: 7
cluster_meta = readRDS(file.path(DATA_DIR, "sci-subset-no-annotation-clusters.rds"))
sci_sub = sci[, sci@meta.data$barcode %in% cluster_meta$barcode]
sci_sub$layer1 = as.factor(cluster_meta$layer1[match(sci_sub@meta.data$barcode, cluster_meta$barcode)])
sci_sub$layer2 = as.factor(cluster_meta$layer2[match(sci_sub@meta.data$barcode, cluster_meta$barcode)])

DimPlot(sci_sub, group.by = c("layer1", "layer2"), ncol = 2, label = TRUE, repel = TRUE) & coord_fixed()
```

```{r}
#| label: marker-dotplot
markers = list(
    Neurons = c("Rbfox3", "Snap25", "Syt1", "Tubb3"),
    Astroependymal = c("Gfap", "Aldh1l1", "Slc1a3", "Sox9"),
    `Immune cells` = c("Cx3cr1", "P2ry12", "Tmem119", "Ptprc"),
    Oligodendrocytes = c("Mbp", "Plp1", "Mog"),
    `Vascular cells` = c("Cldn5", "Pecam1", "Acta2", "Tie1")
)

dot_data = lapply(names(markers), function(ct) {
    dp = DotPlot(sci_sub, features = markers[[ct]], group.by = "layer1")$data
    dp$marker_group = ct
    dp
})
dot_df = do.call(rbind, dot_data)
dot_df$marker_group = factor(dot_df$marker_group, levels = names(markers))
dot_df = dot_df %>%
    group_by(marker_group) %>%
    mutate(pct_scaled = (pct.exp - min(pct.exp)) / (max(pct.exp) - min(pct.exp))) %>%
    ungroup()

ggplot(dot_df, aes(x = features.plot, y = id, size = pct_scaled, colour = avg.exp.scaled)) +
    geom_point() +
    scale_colour_gradient2(low = "steelblue", mid = "white", high = "tomato", name = "Scaled\nexpr") +
    scale_size_continuous(range = c(0, 6), name = "% cells\n(scaled per group)") +
    facet_grid(~ marker_group, scales = "free_x", space = "free_x") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        strip.text = element_text(face = "bold", size = 9),
        panel.spacing = unit(0.3, "lines"),
        panel.border = element_rect(colour = "grey80", fill = NA)
    ) +
    labs(x = NULL, y = "Cluster (layer1)")
```

```{r}
#| label: annotate-clusters
#| fig-width: 10
#| fig-height: 14
layer1_map = c(
    "1" = "Neurons",
    "2" = "Vascular cells",
    "3" = "Astroependymal cells",
    "4" = "Immune cells",
    "5" = "Oligodendrocytes"
)
layer2_map = c(
    "1"  = "OPC",
    "2"  = "Dorsal neurons",
    "3"  = "Vascular cells",
    "4"  = "Ventral neurons",
    "5"  = "Vascular endothelial cells",
    "6"  = "Ependymal cells",
    "7"  = "Astrocytes",
    "8"  = "Peripheral immune cells",
    "9"  = "Microglia",
    "10" = "NFOL",
    "11" = "MOL"
)

sci_sub$cell_type_broad = unname(layer1_map[as.character(sci_sub$layer1)])
sci_sub$cell_type = unname(layer2_map[as.character(sci_sub$layer2)])

DimPlot(sci_sub, group.by = c("cell_type_broad", "cell_type"), ncol = 1, label = TRUE, repel = TRUE, label.size = 3) & coord_fixed()
```

::: {.callout-warning}
## CHECKPOINT
Does the UMAP make biological sense? Which clusters surprise you?

- Are injury-associated cell states separating from homeostatic ones?
- Do you see expected cell types for spinal cord tissue?
- Which cluster would you investigate further and why?
:::

---

## 7. Cell type proportions — Sankey plot

```{r}
#| label: sankey-plot
label_levels = c("uninjured", "young", "old")

prop_df = sci_sub@meta.data %>%
    filter(!is.na(cell_type), label %in% label_levels) %>%
    mutate(label = factor(label, levels = label_levels)) %>%
    group_by(label, cell_type) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(label) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup() %>%
    tidyr::complete(label, cell_type, fill = list(n = 0, prop = 0))

ggplot(prop_df,
       aes(x = label, y = prop, alluvium = cell_type, stratum = cell_type, fill = cell_type)) +
    geom_flow(alpha = 0.7) +
    geom_stratum() +
    theme_minimal() +
    labs(y = "Proportion", x = "Condition", fill = "Cell type")
```

---

## 8. What single-cell resolution gives you

Bulk RNA-seq averages expression across every cell in a tissue sample. A gene upregulated in microglia but downregulated in neurons would appear unchanged — the signals cancel. Single-cell sequencing resolves this by profiling each cell independently.

The Sankey plot above makes this concrete. Cell type **composition** shifts substantially between uninjured and injured spinal cord: immune populations expand, oligodendrocyte lineage cells respond, vascular cells reorganize. None of these compositional changes are visible in a bulk experiment — you see only a noisy average over a moving mixture.

This resolution is what makes Notebook 2 possible. Instead of asking *"which genes change in the injured spinal cord?"*, we can ask *"which genes change specifically in microglia, in astrocytes, in OPCs?"* — questions that bulk DE cannot answer.

```{r}
#| label: save-annotated
saveRDS(sci_sub, file.path(DATA_DIR, "sci-subset-annotated.rds"))
```
