---
title: "Notebook 1 — Statistical Inference & Differential Expression"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(harmony)
library(edgeR)
library(ggplot2)
library(ggalluvial)
library(dplyr)
library(patchwork)
library(clusterProfiler)
library(org.Mm.eg.db)
library(splatter)
library(sparseMatrixStats)

PROJ_DIR <- ""   # set this to your project path, e.g. "/shared/projects/.../projects/C10"
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "figures")
GIT_DIR  <- ""   # set this to your git checkout path, e.g. "~/cajal-project-alan"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

---

## 1. Raw count distributions

We start by looking at the raw count data before any normalisation. Two distributions are informative:

- **Library size per cell**: total UMIs captured — tells you about sequencing depth variation across cells
- **Count values**: the distribution of individual gene counts — tells you about sparsity and the shape of the count distribution

```{r}
#| label: load-raw
sci = readRDS(file.path(DATA_DIR, "sci-subset-atlas-no-annotation.rds"))
counts = GetAssayData(sci, layer = "counts")
```

> **Predict before you run:** What shape do you expect for the library size distribution? Will it be symmetric or skewed?

```{r}
#| label: raw-distributions
#| eval: false
# p1: Library size per cell (total UMI per cell across all genes)
# Hint: colSums(counts) gives total UMI per cell; use scale_x_log10()


# p2: Non-zero count values from 1000 randomly sampled cells
# Hint: sample() cell indices; extract non-zero values via the @x slot of the sparse matrix
# Use set.seed(42) for reproducibility


# Combine p1 + p2 with patchwork

```

**Interpret:** Why do we plot on a log scale? What does the spike of low counts tell you about scRNA-seq data?

---

## 2. Simulating count data

`simulate_counts_naive()` draws counts directly from three statistical distributions. Use it to compare how well each model captures the shape of real count data.

> **Predict before you run:** Which model do you expect to produce the most realistic-looking distribution — Poisson, Gaussian, or NB?

```{r}
#| label: simulate-counts
#| eval: false
# Simulate counts under three models (n_genes = 500, n_cells = 100)
# Available models: "poisson", "gaussian", "nb"


# Combine all non-zero values into a data frame with a 'model' column
# Factor model with levels: Gaussian, Poisson, NB


# Plot as a faceted histogram on log scale with density overlay

```

**Interpret:** Which distribution best matches the real data shape from section 1? What does overdispersion look like compared to Poisson?

---

## 3. Why NBGMM?

To model differential expression in scRNA-seq properly, we need a distribution that matches how count data actually behaves. This section builds intuition for why the **Negative Binomial Generalised Mixed Model (NBGMM)** is the right choice.

### 3a. Overdispersion — why NB over Poisson

**The Poisson assumption.** The simplest count model is Poisson, where the variance equals the mean:

$$\text{Var}[X] = \mathbb{E}[X] = \mu$$

In scRNA-seq, this would mean a gene with mean expression 10 should have variance 10 across cells. In practice, the variance is much higher — this is **overdispersion**.

**Why scRNA-seq is overdispersed.** Gene expression is inherently stochastic. Transcription happens in bursts; cells in the same population differ in their regulatory state, cell cycle phase, and local microenvironment. This biological cell-to-cell variation adds variance on top of the Poisson sampling noise.

**The Negative Binomial distribution** adds a dispersion parameter `size` (also written φ or 1/α):

$$\text{Var}[X] = \mu + \frac{\mu^2}{\text{size}}$$

As `size` → ∞, the NB approaches Poisson. Small `size` (e.g. 0.1–1) means high overdispersion — typical for scRNA-seq.

**What about zero-inflated models?** Single-cell data contains many zeros — often >90% of entries are zero. These zeros come from two sources:

- **Structural zeros**: the gene is genuinely not expressed in that cell
- **Dropout zeros**: the gene was expressed but not captured due to low sequencing depth or PCR failure

A **Zero-Inflated Negative Binomial (ZINB)** model explicitly models this:

$$P(X = 0) = \pi + (1 - \pi) \cdot P_\text{NB}(X = 0)$$

where π is the probability of a structural (dropout) zero. However, in practice NB is often preferred: modern 10x Chromium protocols have lower dropout rates than older protocols, and the two types of zeros are statistically hard to separate. NB already predicts many zeros naturally for low-mean genes, and adding the zero-inflation parameter often increases model complexity without improving DE results.

> **Predict before you run:** On a log-log plot of variance vs. mean, where would you expect Poisson data to fall relative to NB?

```{r}
#| label: demo-overdispersion
#| eval: false
set.seed(42)
n_genes_mv = 500
n_cells_mv = 100
gene_means = exp(rnorm(n_genes_mv, mean = log(5), sd = 1.5))

sim_pois_mv = matrix(rpois(n_genes_mv * n_cells_mv, lambda = gene_means),
    nrow = n_genes_mv, ncol = n_cells_mv)
sim_nb_mv = matrix(rnbinom(n_genes_mv * n_cells_mv, size = 0.5, mu = gene_means),
    nrow = n_genes_mv, ncol = n_cells_mv)

real_means = sparseMatrixStats::rowMeans2(counts)
real_vars = sparseMatrixStats::rowSds(counts)^2

mv_df = rbind(
    data.frame(mean = rowMeans(sim_pois_mv), var = apply(sim_pois_mv, 1, var), model = "Poisson (simulated)"),
    data.frame(mean = rowMeans(sim_nb_mv), var = apply(sim_nb_mv, 1, var), model = "NB (simulated)"),
    data.frame(mean = real_means, var = real_vars, model = "Real data")
)
mv_df = mv_df[mv_df$mean > 0 & mv_df$var > 0, ]
mv_df$model = factor(mv_df$model, levels = c("Poisson (simulated)", "NB (simulated)", "Real data"))

ggplot(mv_df, aes(x = mean, y = var)) +
    geom_point(alpha = 0.4, size = 0.6, colour = "steelblue") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "tomato") +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ model, nrow = 1, scales = "free") +
    labs(
        title = "Mean–variance relationship",
        subtitle = "Real scRNA-seq data is overdispersed like NB — Poisson is constrained to the diagonal",
        x = "Gene mean", y = "Gene variance"
    ) +
    theme_minimal()
```

**Interpret:** The dashed red line is the Poisson expectation (variance = mean). What does it mean for real data points to lie above this line?

### 3b. Replicate structure — why a mixed model

Each biological replicate (animal/sample) has its own baseline expression level — a **random effect** shared by all cells from that sample. When you pool cells across replicates and ignore this structure, the extra between-replicate variance inflates the apparent spread of each group, making it easy to call false positives.

The **mixed model** in NBGMM accounts for this: it explicitly models the within-replicate correlation, so that the test statistic reflects between-replicate variation (the true unit of replication) rather than between-cell variation.

> **Predict before you run:** If each replicate has a random shift in mean expression, what will the pooled distribution look like compared to each individual replicate?

```{r}
#| label: demo-replicates
#| eval: false
set.seed(42)
n_samples_demo = 6
n_cells_demo = 80
base_mean = 5
de_fc_demo = 2

# 3 Control replicates, 3 Treated replicates
group_demo = c(rep("Control", 3), rep("Treated", 3))
group_fc = c(rep(1, 3), rep(de_fc_demo, 3))

# Each replicate has a random multiplicative effect on mean expression
# Hint: use exp(rnorm(...)) to generate log-normal sample effects
sample_effects = ___

# For each replicate, simulate NB counts with mu = base_mean * sample_effect * group_fc
# size = 1 (overdispersed), n_cells_demo cells per replicate
cells_list = lapply(seq_len(n_samples_demo), function(i) {
    mu_i = ___
    data.frame(
        count = rnbinom(n_cells_demo, size = 1, mu = mu_i),
        replicate = paste0("S", i),
        group = group_demo[i]
    )
})
cells_df = do.call(rbind, cells_list)
cells_df$replicate = factor(cells_df$replicate)

# p_pooled: density histogram of count+1, coloured by group, all replicates merged
p_pooled = ___

# p_per_sample: same but faceted by replicate (nrow = 2)
p_per_sample = ___

p_pooled + p_per_sample + plot_layout(widths = c(1, 2))
```

**Interpret:** How does the pooled distribution compare in width to each individual replicate? Why is this a problem if you run a t-test or Wilcoxon test on the pooled cells?

### 3c. DE effect size (`de_fc`)

```{r}
#| label: demo-de-fc
#| eval: false
sim_low_fc = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 1.5, n_samples = 4)
sim_high_fc = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 4, n_samples = 4)

lfc_df = rbind(
    transform(compute_logfc(sim_low_fc), scenario = "de_fc = 1.5"),
    transform(compute_logfc(sim_high_fc), scenario = "de_fc = 4")
)

ggplot(lfc_df, aes(x = logFC, fill = is_de, colour = is_de)) +
    geom_histogram(aes(y = after_stat(density)), bins = 60, position = "identity", alpha = 0.3, colour = "white") +
    geom_density(alpha = 0) +
    facet_wrap(~ scenario, nrow = 1) +
    scale_fill_manual(values = c("grey70", "tomato"), labels = c("not DE", "true DE")) +
    scale_colour_manual(values = c("grey50", "tomato"), labels = c("not DE", "true DE")) +
    labs(x = "log2 fold change (group means)", y = "Density", fill = NULL, colour = NULL) +
    theme_minimal()
```

### 3d. DE probability (`de_prob`)

```{r}
#| label: demo-de-prob
#| eval: false
sim_low_prob = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.05, de_fc = 2, n_samples = 4)
sim_high_prob = simulate_counts(n_genes = 1000, n_cells = 400, de_prob = 0.4, de_fc = 2, n_samples = 4)

de_df = rbind(
    data.frame(scenario = "de_prob = 0.05",
        type = c("DE", "not DE"),
        n = c(length(sim_low_prob$de_genes), 1000 - length(sim_low_prob$de_genes))),
    data.frame(scenario = "de_prob = 0.40",
        type = c("DE", "not DE"),
        n = c(length(sim_high_prob$de_genes), 1000 - length(sim_high_prob$de_genes)))
)

ggplot(de_df, aes(x = scenario, y = n, fill = type)) +
    geom_col() +
    scale_fill_manual(values = c("tomato", "grey70")) +
    labs(y = "Number of genes", fill = NULL) +
    theme_minimal()
```

### 3e. Sequencing depth between replicates (`replicate_effect`)

Real experiments often have systematic differences in sequencing depth across replicates — one sample may have been sequenced more deeply than another, or RNA quality varied. The `replicate_effect` parameter in `simulate_counts()` introduces this by downsampling each replicate's library to a different depth. Setting it to 0 gives perfectly equal depth; larger values exaggerate the variation.

Your task: simulate with no replicate depth effect and with a strong effect, then plot both alongside the real SCI data to judge how realistic each simulation is.

> **Predict before you run:** How would systematic depth differences across replicates affect a DE test that does not account for library size?

```{r}
#| label: replicate-effect
#| eval: false
# Simulate with replicate_effect = 0 and replicate_effect = 0.2
# Use n_genes = 1000, n_cells = 400, de_prob = 0.2, de_fc = 2, n_samples = 4
sim_no_effect = ___
sim_strong_effect = ___

# Build real_lib_df: lib_size = colSums(counts), batch = sci@meta.data$replicate, scenario = "Real data"
real_lib_df = ___

# Order real data replicates by descending median library size
# Hint: group_by(batch) %>% summarise(med = median(lib_size)) %>% arrange(desc(med)) %>% pull(batch)
median_order = ___

# rbind all three scenarios; set scenario and batch as factors with correct level order
lib_df = ___

# Boxplot: x = batch, y = lib_size on log scale, faceted by scenario (nrow = 1, scales = "free")
# No outlier dots; rotate x-axis labels 45 degrees

```

**Interpret:** Does the real data show more variability in library size across replicates than the `replicate_effect = 0` simulation? What might cause this in practice?

---

## 4. QC & normalisation

Standard Seurat QC: remove low-quality cells (too few counts, too many mitochondrial reads) then normalise.

> **Predict before you run:** Which condition do you expect to have higher mitochondrial content — uninjured or injured?

```{r}
#| label: qc-metrics
#| message: false
sci[["pct_mt"]] = PercentageFeatureSet(sci, pattern = "^mt-")

VlnPlot(sci,
    features = c("nFeature_RNA", "nCount_RNA", "pct_mt"),
    group.by = "label",
    pt.size = 0)
```

```{r}
#| label: qc-filter
sci = subset(sci, subset = nCount_RNA > 200 & nCount_RNA < 25000 & pct_mt < 20)
```

```{r}
#| label: normalise
#| message: false
sci = NormalizeData(sci)
sci = FindVariableFeatures(sci, nfeatures = 3000)
sci = ScaleData(sci)
sci = RunPCA(sci, npcs = 30)
```

---

## 5. Harmony integration & UMAP

Replicate effects (batch effects) can dominate the UMAP structure. Harmony corrects for replicate-level variation while preserving biological signal.

> **Predict before you run:** After Harmony, should replicates from the same condition cluster together or be mixed?

```{r}
#| label: harmony-umap
#| eval: false
# Run Harmony integration correcting for "replicate"


# Run UMAP on the Harmony embedding (dims = 1:20)


# Plot UMAP coloured by "label" and "replicate" side by side (ncol = 2)

```

**Interpret:** Does the UMAP show mixing of replicates within conditions? Are conditions (label) still separated?

---

## 6. Cell type annotation

### 6a. Explore clustering resolutions

Different resolutions give different numbers of clusters. Compare a few before settling on one.

```{r}
#| label: clustering
#| message: false
#| eval: false
# Run FindNeighbors using the harmony reduction (dims = 1:20)


# Run FindClusters at resolutions 0.2, 0.5, and 1.0
resolutions = c(0.2, 0.5, 1.0)
for (res in resolutions) {
    # YOUR CODE HERE
}

# Plot all three resolutions side by side using wrap_plots()
# Hint: DimPlot with group.by = paste0("RNA_snn_res.", res)

```

### 6b. Load pre-computed clusters

The instructor has provided pre-computed cluster assignments at two levels of granularity. Each cell has an integer cluster ID — your task is to figure out which integer maps to which cell type using the marker dotplot in 6c.

**Layer 1 — 5 broad cell classes** (integers 1–5, order unknown):

- Neurons
- Vascular cells
- Astroependymal cells
- Immune cells
- Oligodendrocytes

**Layer 2 — 11 fine-grained subtypes** (integers 1–11, order unknown):

- OPC
- Dorsal neurons
- Vascular cells
- Ventral neurons
- Vascular endothelial cells
- Ependymal cells
- Astrocytes
- Peripheral immune cells
- Microglia
- NFOL
- MOL

Use the marker dotplot to determine which integer maps to which cell type.

```{r}
#| label: load-annotation-clusters
cluster_meta = readRDS(file.path(DATA_DIR, "sci-subset-no-annotation-clusters.rds"))
sci_sub = sci[, sci@meta.data$barcode %in% cluster_meta$barcode]
sci_sub$layer1 = as.factor(cluster_meta$layer1[match(sci_sub@meta.data$barcode, cluster_meta$barcode)])
sci_sub$layer2 = as.factor(cluster_meta$layer2[match(sci_sub@meta.data$barcode, cluster_meta$barcode)])

p_clusters = DimPlot(sci_sub, group.by = c("layer1", "layer2"), ncol = 2, label = TRUE, repel = TRUE)
p_clusters
ggsave(file.path(FIG_DIR, "umap_clusters.png"), p_clusters, width = 12, height = 6, dpi = 150)
```

### 6c. Marker dotplot

Choose 3–4 canonical marker genes for each of the 5 layer1 cell types and build a dot plot to identify which integer cluster is which. Dot size = % of cells expressing the gene; colour = scaled expression level.

> **Hint:** Think about well-known markers for neurons (axonal/synaptic proteins), astrocytes (intermediate filaments, glutamate transporters), oligodendrocyte lineage (myelin genes, PDGF receptor), microglia/immune cells (purinergic receptors, complement), and vascular cells (tight junction proteins, endothelial markers). All gene names should be mouse format (e.g. `Mbp`, not `MBP`).

```{r}
#| label: marker-dotplot
#| eval: false
# Define your marker gene lists — one entry per layer1 cell type
markers = list(
    Neurons = c(___),
    Astroependymal = c(___),
    `Immune cells` = c(___),
    Oligodendrocytes = c(___),
    `Vascular cells` = c(___)
)

# Extract DotPlot data per marker group, combine, and scale pct.exp within each group
# Hint: DotPlot(sci_sub, features = markers[[ct]], group.by = "layer1")$data


# Plot using ggplot with facet_grid(~ marker_group, scales = "free_x", space = "free_x")

```

### 6d. Assign cell type labels

Based on the dotplot, fill in the mapping from cluster integer to cell type name.

```{r}
#| label: annotate-clusters
#| eval: false
# Fill in cell type names for each layer1 cluster (integers 1–5)
layer1_map = c(
    "1" = "___",
    "2" = "___",
    "3" = "___",
    "4" = "___",
    "5" = "___"
)

# Fill in cell type names for each layer2 cluster (integers 1–11)
layer2_map = c(
    "1"  = "___",
    "2"  = "___",
    "3"  = "___",
    "4"  = "___",
    "5"  = "___",
    "6"  = "___",
    "7"  = "___",
    "8"  = "___",
    "9"  = "___",
    "10" = "___",
    "11" = "___"
)

sci_sub$cell_type_broad = unname(layer1_map[as.character(sci_sub$layer1)])
sci_sub$cell_type = unname(layer2_map[as.character(sci_sub$layer2)])

p_celltypes = DimPlot(sci_sub, group.by = c("cell_type_broad", "cell_type"), ncol = 2, label = TRUE, repel = TRUE, label.size = 3)
p_celltypes
ggsave(file.path(FIG_DIR, "umap_celltypes.png"), p_celltypes, width = 14, height = 6, dpi = 150)
```

::: {.callout-warning}
## CHECKPOINT
Does the UMAP make biological sense? Which clusters surprise you?

- Are injury-associated cell states separating from homeostatic ones?
- Do you see expected cell types for spinal cord tissue?
- Which cluster would you investigate further and why?
:::

---

## 7. Cell type proportions — Sankey plot

Now that cells are annotated, we can ask: does the **composition** of cell types change between conditions?

> **Predict before you run:** Which cell types do you expect to increase after spinal cord injury? Which might decrease?

```{r}
#| label: sankey-plot
#| eval: false
# Build a proportion data frame:
#   - Filter out NA cell_type values
#   - Group by label and cell_type, count cells, compute proportion within each label
#   - Use tidyr::complete() to fill missing (label, cell_type) combinations with prop = 0
# Hint: label_levels = c("uninjured", "young", "old")


# Plot with ggalluvial:
#   geom_flow() for the ribbons, geom_stratum() for the bars
#   aes(x = label, y = prop, alluvium = cell_type, stratum = cell_type, fill = cell_type)

```

**Interpret:** Which cell type shows the largest proportional shift after injury? Does this match your biological expectations?

---

## 8. What single-cell resolution gives you

Bulk RNA-seq averages expression across every cell in a tissue sample. A gene upregulated in microglia but downregulated in neurons would appear unchanged — the signals cancel. Single-cell sequencing resolves this by profiling each cell independently.

The Sankey plot above makes this concrete. Cell type **composition** shifts substantially between uninjured and injured spinal cord: immune populations expand, oligodendrocyte lineage cells respond, vascular cells reorganize. None of these compositional changes are visible in a bulk experiment — you see only a noisy average over a moving mixture.

This resolution is what makes Notebook 2 possible. Instead of asking *"which genes change in the injured spinal cord?"*, we can ask *"which genes change specifically in microglia, in astrocytes, in OPCs?"* — questions that bulk DE cannot answer.
