---
title: "Notebook 4 (Extended) — Spatial Transcriptomics & Vespucci"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(ggplot2)
library(patchwork)
library(spdep)
library(ranger)
library(pROC)
library(Augur)
library(Vespucci)
library(tidyverse)
library(randomForest)

PROJ_DIR <- "/shared/projects/tp_2630_ubordeaux_neuromics_184418/projects/C14/"   # set this to your project path, e.g. "/shared/projects/.../projects/C10"
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "figures")
GIT_DIR  <- "~/cajal-project-alan"   # set this to your git checkout path, e.g. "~/cajal-project-alan"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

---

Single-cell RNA-seq tells you which genes change in which cell types, but it discards the one thing a tissue has that a dissociated suspension does not: **location**. A neuron in the dorsal horn and one in the ventral horn can express identical genes yet have completely different function and connectivity. Spatial transcriptomics preserves the tissue coordinate system, letting you ask *where* in the spinal cord the injury response is concentrated.

This notebook starts with the standard Seurat Visium workflow to build intuition, then moves to our in-house SCI Visium data. We will see that the standard approach for finding spatially variable genes (Moran's I) has a fundamental blind spot for comparative experiments — and build up to Vespucci as the solution.

---

## 1. Visium basics — mouse brain tutorial

We start with the 10x Genomics mouse brain anterior section, which is the standard Seurat Visium tutorial dataset. This gives us a tissue with clear spatial structure to build intuition before moving to the SCI data.

```{r}
#| label: load-brain
brain = readRDS(file.path(DATA_DIR, "spatial-brain-seurat.rds"))
brain = UpdateSeuratObject(brain)
# SCTransform was already applied before saving — DefaultAssay is "SCT"
```

```{r}
#| label: brain-qc
#| eval: false
# p1: VlnPlot(brain, features="nCount_Spatial", pt.size=0.1) + hide legend
# p2: SpatialFeaturePlot(brain, features="nCount_Spatial") + show legend
# Combine with patchwork: p1 + p2


```

The count distribution across spots is highly variable — spots overlying dense tissue (cortex, hippocampus) tend to capture more RNA. This is a technical effect of local cell density rather than biology.

```{r}
#| label: brain-genes
#| eval: false
# SpatialFeaturePlot(brain, features=c("Hpca", "Ttr"), pt.size.factor=1.6)


```

**Interpret:** Where in the brain do `Hpca` and `Ttr` localise spatially? Do they show tissue-restricted or uniform expression? What kind of genes would you expect Moran's I to pick up?

---

## 2. In-house SCI Visium data

Our SCI Visium data is stored as a standard Seurat object with spatial coordinates in the metadata (`x`, `y`) rather than in the Seurat image slot. We therefore use `ggplot2` directly for all spatial plots.

```{r}
#| label: load-sci-visium
#| eval: false
sci_spatial = readRDS(file.path(DATA_DIR, "sci-spatial.rds"))
cat("Spots:", ncol(sci_spatial), "  Genes:", nrow(sci_spatial), "\n")
cat("Conditions:", paste(unique(sci_spatial$label), collapse = ", "), "\n")

ggplot(sci_spatial@meta.data, aes(x = x, y = y, colour = label)) +
    geom_point(size = 0.6) +
    scale_y_reverse() +
    coord_fixed() +
    facet_wrap(~ label) +
    theme_void() +
    labs(title = "SCI Visium — spot layout by condition") +
    theme(legend.position = "none")
```

```{r}
#| label: sci-normalise
#| eval: false
sci_spatial = NormalizeData(sci_spatial, assay = "RNA", verbose = FALSE)
```

```{r}
#| label: sci-genes-raw
#| eval: false
genes = c("Gfap", "Rbfox3")
expr_mat = as.matrix(GetAssayData(sci_spatial, assay = "RNA", layer = "data")[genes, ])
plot_df = do.call(rbind, lapply(genes, function(g) {
    cbind(sci_spatial@meta.data[, c("x", "y", "label")], expr = expr_mat[g, ], gene = g)
}))

ggplot(plot_df, aes(x = x, y = y, colour = expr)) +
    geom_point(size = 0.4) +
    scale_colour_gradientn(colours = c("grey90", "steelblue", "navy")) +
    scale_y_reverse() +
    coord_fixed() +
    facet_grid(gene ~ label) +
    theme_void() +
    labs(title = "Raw spot expression", colour = "log norm") +
    theme(strip.text = element_text(size = 8))
```

Raw spots are noisy — each spot captures 1–10 cells and the expression is sparse. We can reveal the underlying spatial pattern by fitting a local regression (LOESS) surface across the tissue and predicting onto a regular grid.

```{r}
#| label: sci-genes-loess
#| eval: false
meta = sci_spatial@meta.data
fixed_coords = expand.grid(x = seq(-400, 400, 5), y = seq(-150, 150, 5))

plot_list = list()
for (gene in genes) {
    meta$expr = as.numeric(GetAssayData(sci_spatial, assay = "RNA", layer = "data")[gene, ])
    conditions = unique(meta$label)

    coords = map_dfr(conditions, ~ {
        condition = .x
        fit = loess(expr ~ x * y, data = filter(meta, label == condition), span = 0.02, degree = 1)
        coords = tidyr::crossing(x = seq(-400, 400, 5), y = seq(-150, 150, 5)) %>% mutate(interp = predict(fit, newdata = .)) %>% drop_na(interp) %>%
        mutate(condition = condition)
    })

    range = quantile(coords$interp, c(0.01, 0.99))
    coords$interp = pmax(pmin(coords$interp, range[2]), range[1])
    brks = c(range[1] + 0.1 * diff(range), range[2] - 0.1 * diff(range))
    labels = c("min", "max")

    coords$gene = gene
    p = ggplot(coords, aes(x = x, y = y, fill = interp)) +
        facet_grid(gene ~ condition, switch = "y") +
        geom_raster() +
        scale_y_reverse(expand = c(0, 0), breaks = c(-150, 150), labels = c(expression("R" %->% ""), expression("" %<-% "L")),name = gene) +
        scale_x_continuous(expand = c(0, 0), breaks = c(-400, 400), labels = c(expression("" %<-% "Rostral"), expression("Caudal" %->% ""))) +
        scale_fill_distiller(name = "Expression", palette = "YlOrRd", direction = 1, limits = range, breaks = brks, labels = labels) +
        guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
        coord_fixed() +
        theme_bw() +
        theme(panel.grid = element_blank()) +
        theme(
            axis.title.x = element_blank(),
            axis.title.y = element_blank(),
            axis.text.y = element_text(angle = 90, hjust = c(1, 0)),
            axis.text.x = element_text(hjust = c(0, 1)),
            axis.ticks.x = element_blank(),
            axis.ticks.y = element_blank(),
            axis.ticks.length.x = unit(0, "lines"),
            axis.ticks.length.y = unit(0, "lines"),
            legend.position = "right",
            legend.justification = "bottom",
            legend.key.width = unit(0.18, "lines"),
            legend.key.height = unit(0.18, "lines")
        )
    if (gene != genes[1]) p = p + theme(strip.text.x = element_blank())
    if (gene != genes[length(genes)]) p = p + theme(legend.position = "none")
    plot_list[[gene]] = p
}
wrap_plots(plot_list, ncol = 1)
```

**Interpret:** Which genes show the sharpest spatial boundary in the interpolated surface? Does the pattern shift between conditions? What cell types do you think each gene marks, and what does this tell you about the anatomy of the spinal cord sections?

---

## 3. Cell type deconvolution — RCTD (Demo)

Each Visium spot is ~55 µm in diameter and captures RNA from 1–10 cells. The measured expression is a mixture of whatever cell types happen to sit under that spot. Before interpreting spatial gene expression, we need to know *which cell types are where*.

**RCTD** (Robust Cell Type Decomposition, Cable et al. 2022) solves this by fitting a Poisson model to each spot's counts using our annotated scRNA-seq atlas as a reference. The `spacexr` package is not installed in the course environment so RCTD was run offline; we load the pre-computed proportions here.

```{r}
#| label: rctd-load
#| eval: false
rctd_res = readRDS(file.path(DATA_DIR, "sci-spatial-rctd.rds"))
# Long format: barcode × cell_type × weight × resolution
head(rctd_res)

# Join spatial coordinates
meta_coords = sci_spatial@meta.data[, c("barcode", "x", "y", "label")]
rctd_res = merge(rctd_res, meta_coords, by = "barcode")
```

```{r}
#| label: rctd-plot
#| eval: false
# Plot Immune cells (layer1) as a LOESS-smoothed spatial map
immune = rctd_res[rctd_res$cell_type == "Immune cells" & rctd_res$resolution == "layer1", ]

plot_df_rctd = map_dfr(unique(immune$label), ~ {
    condition = .x
    fit = loess(weight ~ x * y, data = filter(immune, label == condition), span = 0.02, degree = 1)
    tidyr::crossing(x = seq(-400, 400, 5), y = seq(-150, 150, 5)) %>%
        mutate(interp = predict(fit, newdata = .)) %>%
        drop_na(interp) %>%
        mutate(condition = condition)
})

range_rctd = quantile(plot_df_rctd$interp, c(0.01, 0.99))
plot_df_rctd$interp = pmax(pmin(plot_df_rctd$interp, range_rctd[2]), range_rctd[1])
brks_rctd = c(range_rctd[1] + 0.1 * diff(range_rctd), range_rctd[2] - 0.1 * diff(range_rctd))

ggplot(plot_df_rctd, aes(x = x, y = y, fill = interp)) +
    geom_raster() +
    scale_y_reverse(expand = c(0, 0), breaks = c(-150, 150), labels = c(expression("R" %->% ""), expression("" %<-% "L"))) +
    scale_x_continuous(expand = c(0, 0), breaks = c(-400, 400), labels = c(expression("" %<-% "Rostral"), expression("Caudal" %->% ""))) +
    scale_fill_distiller(name = "Weight", palette = "Greys", direction = 1, limits = range_rctd, breaks = brks_rctd, labels = c("min", "max")) +
    guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
    coord_fixed() +
    facet_wrap(~ condition) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text.y = element_text(angle = 90, hjust = c(1, 0)),
          axis.text.x = element_text(hjust = c(0, 1)),
          axis.ticks = element_blank()) +
    labs(title = "Immune cells (RCTD layer1)")
```

**Interpret:** Do the RCTD cell type proportions align with the tissue anatomy — neurons in grey matter, oligodendrocytes in white matter? Does the injured section show a shift in immune cell accumulation compared to uninjured?

**Explore:** Do the RCTD cell type proportions match the known anatomy of spinal cord sections — neurons concentrated in grey matter, oligodendrocytes in white matter, microglia more diffuse? Which spots show the highest immune cell fraction, and does this align with the injury site?

---

## 4. Spatially variable genes — Moran's I

Now that we know *which cells are where*, we want to know *which genes vary spatially*. The standard statistic is **Moran's I**.

For a gene with expression vector **x** across N spots with spatial weight matrix **W**:

$$I = \frac{N}{W_0} \cdot \frac{\mathbf{x}^\top \mathbf{W}\, \mathbf{x}}{\mathbf{x}^\top \mathbf{x}}$$

I = 1: nearby spots are similar (spatial clustering). I = 0: random. I = −1: checkerboard.

```{r}
#| label: moran-brain
#| eval: false
# FindSpatiallyVariableFeatures(brain, features=VariableFeatures(brain)[1:100],
#   selection.method="moransi")
# Print top 10 with SpatiallyVariableFeatures(brain, method="moransi")


```

```{r}
#| label: moran-brain-plot
#| eval: false
# top_svgs: head(SpatiallyVariableFeatures(brain, method="moransi"), 6)
# lapply over top_svgs: SpatialFeaturePlot + labs(title=gene) + no legend
# wrap_plots(ncol=3)


```

**Interpret:** Do the top SVGs match known brain anatomy — do their spatial patterns correspond to recognizable brain structures? What does this tell you about what Moran's I is good at detecting?

---

## 5. The blind spot: condition-specific spatial expression

### The problem

Moran's I measures spatial autocorrelation **within a single tissue section**. It detects genes that cluster in space *within one condition*. But genes that are differentially expressed between conditions only in a specific spatial region — for example, the centre of the spinal cord — have no spatial structure when you examine each condition in isolation. Moran's I is blind to condition-specific spatial patterns.

### Simulation

We use a pre-built simulation: 10,000 cells scattered across a 2D tissue, two conditions (`Perturbation_1`, `Perturbation_2`). A circle of 705 cells in the centre is selectively perturbed — 20% of genes (`truth = 1`) are differentially expressed between conditions **only in this perturbed region**. The remaining 80% of genes (`truth = 0`) are unchanged. 

```{r}
#| label: sim-load
spatial_sim = readRDS(file.path(DATA_DIR, "spatial-simulation-circle.rds"))
spatial_sim_genes = readRDS(file.path(DATA_DIR, "spatial-simulation-circle-de-genes.rds"))
meta_sim = spatial_sim@meta.data
cat("Cells:", ncol(spatial_sim), "  Genes:", nrow(spatial_sim), "\n")
cat("DE genes:", sum(spatial_sim_genes$truth), "  Non-DE:", sum(spatial_sim_genes$truth == 0), "\n")
```

```{r}
#| label: sim-setup
#| eval: false
# p_label: geom_point(shape=21, aes(fill=label)) coloured by label_pal
#   aes(x=y, y=x) — note the axis swap to match reference orientation
# p_circle: map label_sim (background/perturbed) to integer, gradient fill grey90->#2166AC
# p_label | p_circle


```

Now plot the spatial expression of top 3 DE genes (`truth = 1`) by delta mean. We fit a LOESS surface within each condition and predict onto a regular grid — this reveals the circle pattern that is invisible in raw scattered points.

```{r}
#| label: sim-genes
#| eval: false
# Select genes_to_show: top 3 by biggest |delta| and top 3 by smallest (using spatial_sim_genes$delta)
# For each gene: LOESS fit within each label using loess(expr ~ x * y, span=0.02, degree=1)
# Predict onto a regular grid via tidyr::crossing(x=x_seq, y=y_seq) + predict(fit, newdata=.)
# geom_raster() with YlOrRd palette, facet_wrap(~ label)
# Hint: use drop_na(interp) after predict to remove grid points outside the tissue


```

**Interpret:** Do the DE genes (`truth = 1`) show a visible difference between conditions in the centre of the tissue? Do the non-DE genes look similar across conditions? Now ask: would Moran's I be able to find these DE genes?

### Moran's I on the simulation

We run Moran's I across both conditions pooled to find spatially variable genes, then check how well those rankings recover the known DE genes.

```{r}
#| label: sim-moran
#| eval: false
# Subset spatial_sim to Perturbation_1, then sample 2000 cells (set.seed(42))
# Build sparse k=8 NN spatial weight matrix:
#   spdep::knearneigh(coords, k=8) -> knn2nb -> nb2listw(style="W")
# Compute Moran's I for each gene in spatial_sim_genes using spdep::moran()
#   Hint: precompute n_spots and S0 = spdep::Szero(listw_sim) outside apply()
# Build moran_sim_res: data.frame(gene, I), merge with spatial_sim_genes[, c("gene","truth")]
# compute_prc(moran_sim_res$I, moran_sim_res$truth) then auprc()


```

**Interpret:** How does the Moran's I AUPRC compare to the random baseline of 0.5? What does this tell you about whether spatial autocorrelation within one condition can detect genes that are differentially expressed between conditions?

---

## 6. Vespucci — spatial perturbation mapping

### Step 1: The naive approach — Augur at a single barcode

The ideal solution: for each barcode **b**, collect its 50 nearest cells from each condition (by spatial distance) and run `calculate_auc_fast` on those ~100 cells to get a local AUC. Let us see how long that takes for one barcode.

```{r}
#| label: vespucci-one-barcode
#| eval: false
pos_all = as.matrix(meta_sim[, c("x", "y")])

# Build nn_df: for each cell i, find its 100 nearest spatial neighbours with knearneigh()
# Store as data.frame with columns: source_barcode, target_barcode, target_label, neighbor_idx
# Hint: map_dfr over seq_len(nrow(meta_sim)); knearneigh(pos_all, k=100)$nn[i, ] gives neighbours

# Define run_augur_at_barcode(bc, k=50, n_subsamples=10):
#   1. From nn_df, filter to source_barcode==bc, group_by target_label, take top k by neighbor_idx
#   2. Set meta0$cell_type = "vespucci" for those barcodes
#   3. Call Vespucci::calculate_auc_fast(input=counts[, barcodes_keep], meta=meta0,
#        label_col="label", n_subsamples=n_subsamples, select_var=FALSE, n_threads=1)
#   4. Return res$AUC$auc[1]; return NA if fewer than 2 conditions or < 20 cells per condition

# Time one barcode and extrapolate to all barcodes:
bc_example = rownames(meta_sim)[1]
# t0 = proc.time()["elapsed"]; run one barcode; compute t_one; cat estimated hours


```

At this rate, covering all barcodes is computationally infeasible. We need a smarter approach.

---

### Step 2: Build a small training set

Run `calculate_auc_fast` on 10 randomly sampled barcodes. These become our **labelled training examples** — barcodes for which we know the true local AUC.

```{r}
#| label: vespucci-train-augur
#| eval: false
# set.seed(42); sample 10 barcodes from rownames(meta_sim)
# map_dfr: data.frame(barcode=bc, auc=run_augur_at_barcode(bc)), filter(!is.na(auc))
# Print calculated_auc_df


```

---

### Step 3: Distance metrics via `get_distance_metrics`

Instead of running Augur everywhere, Vespucci approximates each barcode's local AUC using cheap pairwise expression statistics. For each barcode, it looks at its k nearest cells from the **opposing condition** and summarises how transcriptomically similar they are — using Pearson and Spearman correlation, each summarised as mean / median / Q1 / Q3 / SD across all cross-condition pairs.

```{r}
#| label: vespucci-distance-metrics
#| eval: false
# Vespucci::get_distance_metrics(
#   input = GetAssayData(spatial_sim, layer="counts"),
#   meta  = data.frame(label=meta_sim$label, barcode=meta_sim$barcode, row.names=rownames(meta_sim)),
#   nn    = nn_df, k=50, label_col="label", barcode_col="barcode",
#   dismay_methods=c("pearson","spearman"), select_var=TRUE, ncores=1
# )
# features = setdiff(colnames(distance_metrics_df), "barcode")
# Print feature names


```

---

### Step 4: Convergence loop

With distance metrics for all barcodes and true AUCs for our 10 training barcodes, we train a `randomForest` meta-classifier. We then sample 10 new barcodes, run `calculate_auc_fast` on each, retrain, and re-predict — repeating until the correlation between successive predictions plateaus.

```{r}
#| label: vespucci-convergence
#| eval: false
# Parameters: barcodes_per_step=10, min_cor=0.99, epsilon=0.01, steps_tracking=3
# While not converged (and remaining barcodes available):
#   1. Sample barcodes_per_step new barcodes, run run_augur_at_barcode on each
#   2. Append to calculated_auc_df; join with distance_metrics_df to get training_data
#   3. randomForest::randomForest(auc ~ ., data=training_data[, c(features,"auc")])
#   4. Predict auc_vals for all barcodes; compute cor() with prev_auc_vals
#   5. Track cor in cor_tracking; increment steps_tracking_count if (r - prev_cor) <= epsilon
#   6. Set converged = TRUE when steps_tracking_count >= steps_tracking && r >= min_cor


```

```{r}
#| label: vespucci-convergence-plot
#| eval: false
# ggplot(cor_tracking[!is.na(cor_tracking$cor), ], aes(x=n_train, y=cor))
# geom_line() + geom_point(size=2)
# Axes: "Training barcodes" vs "Correlation with previous step" (limits c(0,1))


```

Once the correlation plateaus near 1, the meta-classifier has learned a stable mapping from expression statistics to AUC. We take the final predictions as the **spatial perturbation map**.

```{r}
#| label: vespucci-final-map
#| eval: false
# meta_sim$auc_pred = prev_auc_vals matched by barcode to distance_metrics_df$barcode
# LOESS smooth: loess(auc_pred ~ x * y, span=0.015); predict onto meta_sim
# geom_point(aes(x=y, y=x, fill=auc_smooth, color=auc_smooth), shape=21, size=0.3, stroke=0)
# scale_fill/color_distiller(palette="RdYlBu", direction=-1)


```

**Interpret:** Compare this map to the ground truth circle. Does the perturbed region emerge as a spatially coherent high-AUC island? How precisely does Vespucci localise the perturbation signal, and how does this compare to what Moran's I found?

**Explore:** Compare the Vespucci AUC spatial map to the RCTD cell type map. Are the high-AUC regions (most perturbed spots) enriched for particular cell types? What does this tell you about which cell types drive the spatial injury response?

---

### Step 5: Finding DE genes — NBGMM on top barcodes

Once we have the spatial AUC map, `find_spatial_de` uses the **predicted AUC as a continuous label** — rather than selecting a binary top/bottom set of barcodes, every cell contributes and NBGMM tests which genes covary with local perturbation intensity. Replicate is included as a random effect.

```{r}
#| label: vespucci-de
#| eval: false
library(nebula)
# Build auc_df_sim = data.frame(barcode=distance_metrics_df$barcode, auc=prev_auc_vals)
# Vespucci::find_spatial_de(
#   input=GetAssayData(spatial_sim, layer="counts"), meta=meta_sim,
#   auc_df=auc_df_sim, barcode_col="barcode", replicate_col="replicate"
# )
# Print head(de_sim, 10)


```

```{r}
#| label: vespucci-de-plot
#| eval: false
# top_de_genes: head(de_sim$feature[de_sim$p_val_adj < 0.05], 6)
# For each gene: LOESS surface on x/y within each label (span=0.02)
# tidyr::crossing(x=x_seq_de, y=y_seq_de) grid + predict + drop_na
# geom_raster + YlOrRd palette + facet_wrap(~ label) + aes(x=y, y=x)
# wrap_plots(de_plots, ncol=2)


```

---

### A note on data independence: countsplitting

There is a subtle statistical issue: we used the full expression matrix both to compute the spatial AUC map (step 3, distance metrics) **and** to find DE genes (step 5, nebula). This is **data leakage** — the DE test is not independent of the spatial selection step, inflating false positives.

The solution is **countsplitting** (Neufeld et al. 2022): split each cell's raw UMI counts into two independent halves (exploiting the Poisson splitting property). Use one half to compute the AUC map; use the other half for the DE test. The two halves are statistically independent, so the DE test is valid.

---

## 7. Apply Vespucci to the SCI atlas

```{r}
#| label: vespucci-sci
#| eval: false
sci_spatial = readRDS(file.path(DATA_DIR, "sci-spatial-mapped.rds"))
counts_sc = GetAssayData(sci_spatial, layer = "counts")
meta_sc = sci_spatial@meta.data

vespucci_res = Vespucci::calculate_auc_fast(
    input = counts_sc,
    meta = meta_sc,
    label_col = "label",
    cell_type_col = "barcode"
)
auc_scores = setNames(vespucci_res$AUC$auc, vespucci_res$AUC$cell_type)
meta_sc$auc_pred = auc_scores[match(meta_sc$barcode, names(auc_scores))]

fit_sc = loess(auc_pred ~ x * y, data = meta_sc, span = 0.015)
meta_sc$auc_smooth = predict(fit_sc, meta_sc)
rng_sc = range(meta_sc$auc_smooth, na.rm = TRUE)
brks_sc = c(rng_sc[1] + 0.1 * diff(rng_sc), rng_sc[2] - 0.1 * diff(rng_sc))

ggplot(meta_sc[!is.na(meta_sc$auc_smooth), ],
       aes(x = x, y = y, fill = auc_smooth, color = auc_smooth)) +
    geom_point(size = 0.3, shape = 21, stroke = 0) +
    scale_fill_distiller(name = "AUC", palette = "RdYlBu", direction = -1,limits = rng_sc, breaks = brks_sc, labels = format(rng_sc, digits = 2)) +
    scale_color_distiller(name = "AUC", palette = "RdYlBu", direction = -1, limits = rng_sc, breaks = brks_sc, labels = format(rng_sc, digits = 2)) +
    guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE),
           color = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
    scale_y_reverse() + coord_fixed() + facet_wrap(~ label) + theme_bw() +
    theme(panel.grid = element_blank(), axis.title = element_blank(),
          axis.text = element_blank(), axis.ticks = element_blank(),
          legend.position = "right", legend.justification = "bottom",
          legend.key.width = unit(0.18, "lines"), legend.key.height = unit(0.18, "lines")) +
    labs(title = "Vespucci -- spatial perturbation map (SCI)")
```

```{r}
#| label: vespucci-sci-de
#| eval: false
library(nebula)

auc_df_sc = data.frame(barcode = meta_sc$barcode, auc = meta_sc$auc_pred)

de_sc = Vespucci::find_spatial_de(
    input = counts_sc,
    meta = meta_sc,
    auc_df = auc_df_sc,
    barcode_col = "barcode",
    replicate_col = "replicate"
)
head(de_sc, 10)
```

```{r}
#| label: vespucci-sci-de-plot
#| eval: false
sci_spatial = NormalizeData(sci_spatial, assay = "RNA", verbose = FALSE)
top_de_sc = head(de_sc$feature[de_sc$p_val_adj < 0.05], 6)
meta_sc_plot = meta_sc[, c("x", "y", "label")]

x_seq_sc = seq(floor(min(meta_sc_plot$x)), ceiling(max(meta_sc_plot$x)), length.out = 80)
y_seq_sc = seq(floor(min(meta_sc_plot$y)), ceiling(max(meta_sc_plot$y)), length.out = 80)

norm_sc = as.matrix(GetAssayData(sci_spatial, assay = "RNA", layer = "data"))

de_sc_plots = lapply(top_de_sc, function(gene) {
    meta_sc_plot$expr = norm_sc[gene, rownames(meta_sc_plot)]
    plot_df = map_dfr(unique(meta_sc_plot$label), ~ {
        fit = loess(expr ~ x * y, data = filter(meta_sc_plot, label == .x),
                    span = 0.02, degree = 1)
        tidyr::crossing(x = x_seq_sc, y = y_seq_sc) %>%
            mutate(interp = predict(fit, newdata = .), label = .x) %>%
            drop_na(interp)
    })
    rng_g = quantile(plot_df$interp, c(0.01, 0.99), na.rm = TRUE)
    plot_df$interp = pmax(pmin(plot_df$interp, rng_g[2]), rng_g[1])
    brks_g = c(rng_g[1] + 0.1 * diff(rng_g), rng_g[2] - 0.1 * diff(rng_g))
    ggplot(plot_df, aes(x = x, y = y, fill = interp)) +
        ggtitle(gene) +
        geom_raster() +
        scale_fill_distiller(name = "Expression", palette = "YlOrRd", direction = 1,limits = rng_g, breaks = brks_g, labels = c("min", "max")) +
        guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
        scale_x_continuous(expand = c(0, 0)) + scale_y_reverse(expand = c(0, 0)) +
        coord_fixed() + theme_bw() + theme(panel.grid = element_blank(),
              axis.title = element_blank(), axis.text = element_blank(),
              axis.ticks = element_blank()) +
        facet_wrap(~ label)
})
wrap_plots(de_sc_plots, ncol = 2)
```

```{r}
#| label: vespucci-sci-clusters
#| eval: false
# Cluster DE genes by their spatial expression pattern
sig_genes = de_sc$feature[de_sc$p_val_adj < 0.05]

genes_cluster_df = Vespucci::find_de_clusters(
    input = norm_sc,
    meta = meta_sc,
    de_genes = sig_genes,
    barcode_col = "barcode",
    norm = FALSE
)
table(genes_cluster_df$cluster)
```

```{r}
#| label: vespucci-sci-cluster-plot
#| eval: false
# Spatial map: mean expression of each gene cluster per spot
cluster_ids = sort(unique(genes_cluster_df$cluster))

cluster_spatial = map_dfr(cluster_ids, function(cl) {
    cl_genes = genes_cluster_df$gene[genes_cluster_df$cluster == cl]
    mean_expr = colMeans(norm_sc[cl_genes, meta_sc$barcode, drop = FALSE])
    data.frame(barcode = names(mean_expr), mean_expr = mean_expr, cluster = paste0("Cluster ", cl))
}) %>% left_join(meta_sc[, c("barcode", "x", "y")], by = "barcode")

cluster_plots = lapply(unique(cluster_spatial$cluster), function(cl) {
    df = filter(cluster_spatial, cluster == cl)
    fit = loess(mean_expr ~ x * y, data = df, span = 0.015)
    df$smooth = predict(fit, df)
    rng = range(df$smooth, na.rm = TRUE)
    brks = c(rng[1] + 0.1 * diff(rng), rng[2] - 0.1 * diff(rng))
    ggplot(df[!is.na(df$smooth), ], aes(x = x, y = y, fill = smooth, color = smooth)) +
        ggtitle(cl) +
        geom_point(size = 0.3, shape = 21, stroke = 0) +
        scale_fill_distiller(name = "Mean expr", palette = "YlOrRd", direction = 1,limits = rng, breaks = brks, labels = c("low", "high")) +
        scale_color_distiller(name = "Mean expr", palette = "YlOrRd", direction = 1,limits = rng, breaks = brks, labels = c("low", "high")) +
        guides(fill = guide_colorbar(frame.colour = "black", ticks = FALSE),
               color = guide_colorbar(frame.colour = "black", ticks = FALSE)) +
        scale_y_reverse() + coord_fixed() + theme_bw() +
        theme(panel.grid = element_blank(), axis.title = element_blank(),
              axis.text = element_blank(), axis.ticks = element_blank())
})
wrap_plots(cluster_plots)
```

```{r}
#| label: vespucci-sci-gsea
#| eval: false
library(clusterProfiler)
library(org.Mm.eg.db)

gsea_results = lapply(cluster_ids, function(cl) {
    cl_genes = genes_cluster_df$gene[genes_cluster_df$cluster == cl]
    tryCatch(
        enrichGO(gene = cl_genes, OrgDb = org.Mm.eg.db, keyType = "SYMBOL", ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.05, qvalueCutoff = 0.1),
        error = function(e) NULL
    )
})
names(gsea_results) = paste0("Cluster ", cluster_ids)

# Top 5 terms per cluster
gsea_df = map_dfr(names(gsea_results), function(cl_name) {
    res = gsea_results[[cl_name]]
    if (!is.null(res) && nrow(res) > 0)
        head(as.data.frame(res), 5) %>% mutate(cluster = cl_name)
})

ggplot(gsea_df, aes(x = reorder(Description, -log10(p.adjust)), y = -log10(p.adjust), fill = cluster)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    facet_wrap(~ cluster, scales = "free_y") +
    scale_x_discrete(labels = function(x) str_wrap(x, 30)) +
    theme_bw() + theme(panel.grid.minor = element_blank(), axis.title.y = element_blank()) +
    labs(y = "-log10 adjusted p-value", title = "GO:BP enrichment per DE gene cluster")
```

---

::: {.callout-warning}
## CHECKPOINT

What does spatial location add that single-cell alone could not tell you?

- From the simulation: DE genes (`truth = 1`) are upregulated between conditions only in the perturbed circle. Was Moran's I AUPRC above the random baseline of 0.5? What does this tell you about whether SVG analysis is an appropriate tool for this kind of comparison?
- Look at the Vespucci spatial AUC map: does the region of high AUC match the ground truth circle? How precisely does Vespucci localise the perturbation signal?
- Looking at the Vespucci AUC map for the SCI data: is the injury response uniform across the spinal cord cross-section, or concentrated in specific anatomical regions? How does this relate to the RCTD cell type distribution?
- Moran's I is applied within a single condition. Vespucci classifies *between* conditions at each location. What does this mean for experimental design — can you run Vespucci on a single tissue section?
:::

---

## Challenge: Apply to your own data

Find a publicly available **spatial transcriptomics dataset** and apply the spatial analysis methods from this notebook. A single tissue section is sufficient for Moran's I; two or more sections with different conditions are needed to run Vespucci.

### Where to find data

- **10x Genomics public datasets** — search "Visium" at `10xgenomics.com/datasets`; processed outputs are available for mouse brain, human brain, mouse spinal cord, and more
- **Zenodo / GEO** — many spatial papers deposit SpaceRanger outputs or Seurat `.rds` objects; load with `Load10X_Spatial()` or `readRDS()`
- **`SpatialExperiment`** Bioconductor package — small example datasets installable in R

### What your dataset needs

| Data | Content |
|---|---|
| Spatial coordinates | `x`, `y` per spot |
| Expression matrix | Raw counts (for `SCTransform` and Moran's I) |
| `label` *(optional)* | Condition per spot — only needed for Vespucci |
| `replicate` *(optional)* | Replicate ID — only needed for Vespucci DE |

### Tasks

1. Load the spatial data and visualise the tissue layout using the spatial coordinates.
2. Run `SCTransform` and identify variable features.
3. Run `FindSpatiallyVariableFeatures` with `selection.method = "moransi"`. What are the top spatially variable genes?
4. Do the top genes correspond to known anatomical gradients, cell type markers, or tissue layers?
5. *(If 2+ conditions)* Run `Vespucci::run_vespucci()` and produce a spatial AUC map. Where is the perturbation most concentrated?

```{r}
#| label: byod-nb4
#| eval: false
# Load a Visium dataset
# spatial_obj = Load10X_Spatial(data.dir = file.path(PROJ_DIR, "your-visium-dir/"))
# OR: spatial_obj = readRDS(file.path(PROJ_DIR, "your-spatial.rds"))

# Normalise and find SVGs
# spatial_obj = SCTransform(spatial_obj, assay = "Spatial", verbose = FALSE)
# spatial_obj = FindVariableFeatures(spatial_obj)
# spatial_obj = FindSpatiallyVariableFeatures(spatial_obj,
#     features = VariableFeatures(spatial_obj),
#     selection.method = "moransi")

# Extract Moran's I results
# svf_meta = spatial_obj@assays[["SCT"]]@meta.features
# moran_res = data.frame(gene = rownames(svf_meta), I = svf_meta$MoransI_observed)
# head(moran_res[order(-moran_res$I, na.last = TRUE), ], 20)

# Vespucci (if 2+ conditions, with label and replicate columns in meta.data)
# vespucci_res = Vespucci::run_vespucci(
#     input = GetAssayData(spatial_obj, layer = "counts"),
#     meta  = spatial_obj@meta.data
# )
```

**What did you find?** List the top 5 spatially variable genes. Do they correspond to known spatial gradients or cell type markers in your tissue?
