---
title: "Notebook 3 — Cell Population Prioritization with Augur"
subtitle: "CAJAL Neuromics 2026 · Spinal Cord Injury"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(Seurat)
library(ggplot2)
library(patchwork)
library(sparseMatrixStats)
library(Augur)

PROJ_DIR <- "/shared/projects/tp_2630_ubordeaux_neuromics_184418/projects/C14/"   # set this to your project path, e.g. "/shared/projects/.../projects/C10"
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "figures")
GIT_DIR  <- "~/cajal-project-alan"   # set this to your git checkout path, e.g. "~/cajal-project-alan"
source(file.path(GIT_DIR, "setup-quarto-visualization.R"))
```

---

## Cell population prioritization

In Notebook 2, you identified differentially expressed genes per cell type. A natural next question: **which cell type is most perturbed by spinal cord injury?**

Counting significant DE genes per cell type is tempting but misleading — it is biased by cell count, effect size, and choice of significance threshold. A more principled question is: *how distinguishable are the two conditions from each other within each cell type?* Train a classifier to predict condition from single-cell expression and evaluate with AUC — high AUC means the cell type's transcriptome is separable by condition, i.e. perturbed. AUC = 0.5 means conditions are indistinguishable. This is the core idea of **Augur**, which repeatedly subsamples a small balanced set of cells and averages classifier AUC across trials so that cell types are compared on equal footing regardless of size.

---

## 1. Cell population prioritization on the SCI atlas

```{r}
#| label: load-sci
#| eval: false
# Load the real sci-subset-atlas-annotated.rds and print cell/condition counts
# Hint: sci = readRDS(file.path(DATA_DIR, "sci-subset-atlas-annotated.rds")); table(sci$label)


```

Run Augur at three annotation resolutions across all pairwise condition comparisons.

```{r}
#| label: augur-sci
#| eval: false
# Nested loop: comparisons (young vs uninjured, old vs uninjured, young vs old)
#   × layers (layer1, layer2, layer3)
# Subset sci to the two conditions, then call:
#   Augur::calculate_auc(sci_sub, label_col="label", cell_type_col=layer, n_threads=4)
# Store each result in sci_results[[key]]


```

```{r}
#| label: augur-sci-plot
#| eval: false
# lapply over sci_results: plot_lollipop(res) + labs(title=key)
# wrap_plots(plots, ncol=3)


```

**Interpret:** Do the same cell types rank highly across all three comparisons, or does perturbation pattern differ between young and old injury? Does layer3 (subtypes) reveal structure hidden at layer1?

**Explore:** Looking at the Augur results across all three comparisons (young vs uninjured, old vs uninjured, young vs old), which comparison shows the most widespread perturbation? Does this match your expectation about how aging interacts with injury?

---

::: {.callout-warning}
## CHECKPOINT

When does prioritization tell you something DE cannot?

- Do the same cell types rank highly across all three comparisons? Does layer resolution change the picture?
- Can you think of a real biological scenario where a cell type has many DE genes but low Augur AUC? Or few DE genes but high AUC?
:::
